### 2020-04-07 src/paper.tex
first template setup

% COMSYS Seminar/Preseminar Paper Template
% ----------------------------------------
% Based on the original IEEE template package; see files README, changelog.txt for details and copyright information.

\documentclass[conference]{IEEEtran}

%\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Multithreaded Debugging}

\author{\IEEEauthorblockN{Marvin Alexander Strangfeld}
\IEEEauthorblockA{
    \textit{RWTH Aachen University}\\
    marvin.strangfeld@rwth-aachen.de}
}

\maketitle

\begin{abstract}
\end{abstract}

\begin{IEEEkeywords}
\end{IEEEkeywords}

\section{Introduction}
Test \cite{acm2002}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}

### 2020-05-04 examples/deadlock/main.go
add real code examples for deadlock

@@ -1,0 +1,23 @@
+package main
+
+import (
+	"fmt"
+	"sync"
+)
+
+var data = []string{"Earth", "Mars", "Venus", "Saturn"}
+
+// Based on "Understanding real-worldconcurrency bugs in go"
+// by T. Tu, X. Liu, L. Song and Y. Zhang
+func main() {
+	var group sync.WaitGroup
+	group.Add(len(data))
+	for _, d := range data {
+		go func(d string) {
+			fmt.Printf("Processing: %s\n", d)
+			defer group.Done()
+		}(d)
+		group.Wait() // This is causing the deadlock
+	}
+	// group.Wait() // Move here to fix it
+}

### 2020-05-06 src/paper.tex
first draft for a paper

@@
-% COMSYS Seminar/Preseminar Paper Template
-% ----------------------------------------
-% Based on the original IEEE template package; see files README, changelog.txt for details and copyright information.
-
+% ------------------------------------ %
+% Copyright (C) 2020 Marvin Strangfeld %
+% ------------------------------------ %
 \documentclass[conference]{IEEEtran}
 
-%\IEEEoverridecommandlockouts
 \usepackage{cite}
@@
 \usepackage{textcomp}
+\usepackage{listings}
+\usepackage{xcolor}
+\usepackage[hidelinks]{hyperref}
+\usepackage{cleveref}
+
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{1,1,1}
+
+\lstdefinestyle{mystyle}{
+    backgroundcolor=\color{backcolour},
+    commentstyle=\color{codegreen},
+    keywordstyle=\color{magenta},
+    numberstyle=\tiny\color{codegray},
+    stringstyle=\color{codepurple},
+    basicstyle=\ttfamily\footnotesize,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=b,
+    keepspaces=true,
+    numbers=left,
+    numbersep=5pt,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2
+}
+
+\lstset{style=mystyle}
+
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 
-\title{Multithreaded Debugging}
+\title{Finding Concurrency Bugs in Production}
 
@@
 \begin{abstract}
+% TODO: Why in production?
+Today's computer architectures rely more and more on parallel computing.
+Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
+To utilize the full performance of a device, programmers need to write their software with a high degree of parallelism.
+This introduces the potential for bugs that would not occur in sequential programs due to non-determinism of the thread scheduler.
+For this purpose tools and languages have been developed to ease the creation of concurrency aware programs.
+One language this paper will explicitly analyze is \emph{Go} because it claims to be parallel by design and invented to ease the creation of multi-threaded applications.
+In this paper we will have a look at the tools that already exist today to find, reproduce and fix concurrency bugs in real-world software that runs in.
+We will mainly focus on the possibilities to dynamically detect so called ``Heisenbugs'' and how to efficiently fix them by using automated to semi-automated techniques.
 \end{abstract}
@@
 \begin{IEEEkeywords}
+    Debugging, Software Engineering
 \end{IEEEkeywords}
 
+% ------------------------------------ %
+% ----------- INTRODUCTION ----------- %
+% ------------------------------------ %
 \section{Introduction}
-Test \cite{acm2002}
+\begin{quote}
+``Everyone knows that debugging is twice as hard as writing a program in the first place.
+So if you're as clever as you can be when you write it, how will you ever debug it?'' --- Brian W.
+Kernighan\cite{kernighan1974elements}
+\end{quote}
+
+Writing concurrent programs can be very challenging.
+Computational work gets distributed among different processes and threads that need to synchronize to be able to speed up execution time on multi-core architectures.
+Due to this complexity it is no surprise that bugs that are correlated to concurrency and parallel program execution occur quite often.
+These bugs are also the most difficult to debug because they are so complex in their nature and not easy to reproduce.\cite{tu2019go}
+In 1986 Jim Gray introduced the term ``Heisenbug'' for these kind of software failures.
+He wrote:
+
+\begin{quote}
+``The assertion that most production software bugs are soft -- Heisenbugs that go away when you look at them -- is well known to systems programmers.
+Bohrbugs, like the Bohr atom, are solid, easily detected by standard techniques, and hence boring.
+But Heisenbugs may elude a bugcatcher for years of execution.
+Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.
+This is analogous to the Heisenberg Uncertainty Principle in Physics.''\cite{gray1986computers}
+\end{quote}
+
+The complexity and uncertainty of these bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to fix, detect or even avoid them.
+
+In this paper we will evaluate what kind of tools and techniques have already been developed and can be used to ease the process of debugging.
+A main focus lies on detecting bugs efficiently in production.
+Production in this case means that developers should be able to fix bugs effectively in a minimum amount of time.
+Also, the occurrence of concurrency bugs should be reported immediately with a minimum overhead of work and computational resources.
+
+% ------------------------------------ %
+% --- TAXONOMY OF CONCURRENCY BUGS --- %
+% ------------------------------------ %
+\section{Taxonomy of Concurrency Bugs}
+\begin{figure}
+    \includegraphics[width=\linewidth]{figures/ConcurrencyBugClasses.pdf}
+    \caption{A Taxonomy of Concurrency Bugs -- based on\cite{tchamgoue2012testing}}
+    \label{fig:classes}
+\end{figure}
+
+\Cref{fig:classes} shows a taxonomy of concurrency bugs based on the work of Tchamgoue, Kim and Jun. \cite{tchamgoue2012testing}
+For this paper we also distinguish between the two main categories of concurrency bugs: \emph{Blocking} and \emph{non-blocking}.
+Blocking bugs are bugs where the execution of a program is unintentionally blocked and the program can not terminate.
+Non-blocking bugs are often harder to find because they can occur even when the termination of a program was successful but the result is wrong.
+
+\subsection{Deadlocks}
+\begin{lstlisting}[language=Go, label=lst:deadlock, caption=Deadlock example -- based on \cite{tu2019go}]
+var group sync.WaitGroup
+group.Add(len(data))
+for _, d := range data {
+    go func(d string) {
+        fmt.Printf("Processing: %s\n", d)
+        defer group.Done()
+    }(d)
+    group.Wait() // Causing the bug
+}
+\end{lstlisting}
+
+One example for blocking bugs are \emph{deadlocks} where circular dependencies between resources block the flow of a program.
+\Cref{lst:deadlock} shows one example of such a deadlock in a Go program.
+The problem is a \emph{blocking synchronization} where the \lstinline{group.Wait()} inside the for-loop is causing the block.
+Another common problem are \emph{blocking operations} that cause a thread to return before it could finish or redirecting it to another part of the application.
+
+\subsection{Data Races}
+\begin{lstlisting}[language=Go, caption=Data race example -- based on \cite{goRaceDetector}]
+func main() {
+	c := make(chan bool)
+	m := make(map[string]string)
+	go func() {
+		m["1"] = "a" // First conflicting access.
+		c <- true
+	}()
+	m["2"] = "b" // Second conflicting access.
+	<-c
+}
+\end{lstlisting}
+
+\subsection{Atomicity and Order Violations}
+``Around one third of the examined non-deadlock concurrency bugs are caused by violation to programmers' order intentions, which may not be easily expressed via synchronization primitives like locks and transactional memories''\cite{lu2008mistakes}
+
+% ------------------------------------ %
+% ---- CONCURRENCY-AWARE TESTING ----- %
+% ------------------------------------ %
+\section{Concurrency-aware Testing}
+Extensive testing has proven to minimize the bugs of software in production.\cite{makinen2014testing}
+However, traditional testing mostly covers sequential errors and can not detect concurrency bugs effectively.\cite{lu2008mistakes}
+
+% ------------------------------------ %
+% ------ STATIC CODE ANALYSIS -------- %
+% ------------------------------------ %
+\section{Static Code Analysis}
+
+\section{Dynamic Code Analysis}
+
+\section{Avoiding Concurrency Bugs by Language Design}
+% TODO: Go + STM as outlook for
+% Go is not as good as it seems but STM seems promising
+Go is a statically typed and natively compiled programming language developed by Google Inc.
+It gained a lot of popularity after it's first stable release in 2009.
+Today widely used projects like Docker, Kubernetes and Prometheus are build using Go.
+In the official documentation of Go it claims: ``Go is expressive, concise, clean, and efficient.
+Its concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines, while its novel type system enables flexible and modular program construction.''~\footnote{https://golang.org/doc/}
+So, Go tries to tackle some of the problems that arise when writing concurrent software application.
+Projects written in Go tend to have more concurrency than projects written on traditional programming languages such as C.\cite{tu2019go}
+Although Go encourages programmers to use concurrency by message passing, it is also possible to make use of shared memory.
+
+\section{Conclusion}
 

### 2020-05-08 src/paper.tex
Add more references and use final structure

--- 2020-04-07 src/paper.tex
@@
 \usepackage{textcomp}
+\usepackage{listings}
+\usepackage{xcolor}
+\usepackage[hidelinks]{hyperref}
+\usepackage{cleveref}
+\usepackage{float}
+
+% TODO: Correctly align the caption with the listing frame
+\usepackage{caption}
+\DeclareCaptionFont{white}{\color{white}}
+\DeclareCaptionFormat{listing}{\colorbox{darkgray}{\parbox{\dimexpr\linewidth-0.7em\relax}{#1#2#3}}}
+\captionsetup[lstlisting]{format=listing,labelfont=white,textfont=white}
+
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{1,1,1}
+
+\lstdefinestyle{mystyle}{
+    backgroundcolor=\color{backcolour},
+    commentstyle=\color{codegreen},
+    keywordstyle=\color{magenta},
+    numberstyle=\tiny\color{codegray},
+    stringstyle=\color{codepurple},
+    basicstyle=\ttfamily\footnotesize,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=t,
+    keepspaces=true,
+    numbers=left,
+    numbersep=5pt,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2,
+    frame=single,
+    xleftmargin=2em,
+    framexleftmargin=1.57em,
+    framexrightmargin=-0.5em
+}
+
+\lstset{style=mystyle}
+\renewcommand{\lstlistingname}{Code}
+\Crefname{listing}{Code}{Codes}
+\crefname{listing}{code}{codes}
+
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 \begin{abstract}
+% TODO: Why in production?
+Today's computer architectures rely more and more on parallel computing.
+Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
+To utilize the full performance of a device, programmers need to write their software with a high degree of parallelism.
+This introduces the potential for bugs that would not occur in sequential programs due to non-determinism of the order of execution.
+For this purpose tools and languages have been developed to ease the creation of concurrency aware programs and help programmers find concurrency bugs.
+In this paper we will evaluate the different tools and techniques that are available to find, reproduce and fix concurrency bugs from the viewpoint of how usable they are in a production environment.
 \end{abstract}
@@
 \begin{IEEEkeywords}
+    [Software Engineering]: Testing and Debugging
 \end{IEEEkeywords}
 
+% ------------------------------------ %
+% ----------- INTRODUCTION ----------- %
+% ------------------------------------ %
 \section{Introduction}
-Test \cite{acm2002}
+
+\subsection{Heisenbugs}
+
+\begin{quote}
+``Everyone knows that debugging is twice as hard as writing a program in the first place.
+So if you're as clever as you can be when you write it, how will you ever debug it?'' --- Brian W.
+Kernighan\cite{kernighan1974elements}
+\end{quote}
+
+Writing concurrent programs can be very challenging.
+Computational work gets distributed among different processes and threads that need to synchronize to be able to speed up execution time on multi-core architectures.
+Due to this complexity it is no surprise that bugs that are correlated to concurrency and parallel program execution occur quite often.
+These bugs are also the most difficult to debug because they are so complex in their nature and not easy to reproduce.~\cite{tu2019go}
+This is because of the non-determinism that is introduced when multiple \emph{threads} or \emph{processes} are executed simultaneously.
+In contrast to sequential programs that only run on one thread and process, concurrent programs have multiple paths of execution that are executed independently from each other.
+The order these threads are executed is determined by the operating system's scheduler.
+Because a programmer like the program itself can not predict what resources also need processing time on a machine, the choice of the schedule has to be handled as non-predictable and therefore non-deterministic.
+But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread executions appear can not be reliably reproduced without knowing the schedule it appeared.
+
+In 1986 Jim Gray introduced the term ``Heisenbug'' for these kind of software failures.
+He wrote:
+
+\begin{quote}
+``The assertion that most production software bugs are soft -- Heisenbugs that go away when you look at them -- is well known to systems programmers.
+Bohrbugs, like the Bohr atom, are solid, easily detected by standard techniques, and hence boring.
+But Heisenbugs may elude a bugcatcher for years of execution.
+Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.
+This is analogous to the Heisenberg Uncertainty Principle in Physics.''\cite{gray1986computers}
+\end{quote}
+
+The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix or even avoid them.
+Thus, numerous tools and techniques have been developed that propose different solutions to the problem of concurrency bugs.
+
+In this paper we will evaluate some of these and how they can be used to ease the process of debugging concurrent bugs.
+The main focus is on detecting bugs efficiently in production, which means that developers should be able to fix bugs with a minimum amount of work and time.
+Also, the occurrence of concurrency bugs should be reported immediately with minimal overhead of computational resources.
+
+All code examples are written in Go which is a statically typed programming language developed by Google.
+The reason we are using Go and not C/C++ or Java is the recent popularity of it for writing highly concurrent programs.
+Popular open-source projects like \emph{Kubernetes}, \emph{Docker} or \emph{Prometheus} are all written in Go.
+In the official documentation of Go it claims: ``Go is expressive, concise, clean, and efficient.
+Its concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines, while its novel type system enables flexible and modular program construction.''~\footnote{\url{https://golang.org/doc/}}
+So, Go tries to tackle some of the problems that arise when writing concurrent software application by their design of language.
+% TODO: How does Go want to archive this?
+Although it encourages programmers to use concurrency by \emph{message passing}, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
+Projects written in Go also tend to have more concurrency than projects written in traditional programming languages such as C~\cite{tu2019go} what makes it even more important for developers to catch concurrency bugs effectively.
+
+\subsection{Table of Contents}
+
+\Cref{sct:taxonomy} will give a brief introduction on the different types of concurrency bugs and their causes.
+\Cref{sct:dynamic} wil cover some techniques to dynamically find concurrency bugs and reliably reproduce them.
+\Cref{sct:testing} will show different methods of concurrency-aware testing to detect concurrency bugs automatically by manipulating the thread scheduler.
+And \Cref{sct:static} will finally cover how to detect concurrency bugs with static code analysis.
+
+% ------------------------------------ %
+% --- TAXONOMY OF CONCURRENCY BUGS --- %
+% ------------------------------------ %
+\section{Taxonomy of Concurrency Bugs}
+\label{sct:taxonomy}
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figures/ConcurrencyBugClasses.pdf}
+    \caption{A Taxonomy of Concurrency Bugs -- based on\cite{tchamgoue2012testing}}
+    \label{fig:classes}
+\end{figure}
+
+\Cref{fig:classes} shows a taxonomy of concurrency bugs based on the work of Tchamgoue, Kim and Jun.~\cite{tchamgoue2012testing}
+For this paper we also distinguish between the two main categories of concurrency bugs: \emph{Blocking} and \emph{non-blocking}.
+Blocking bugs are bugs where the execution of a program is unintentionally blocked and the program can not terminate.
+The manifestation of these kind of bugs is very noticeable because the whole application freezes and has to be shutdown before the computation can continue.
+Non-blocking bugs are often harder to find because they can occur even when the termination of a program was successful but the result is wrong.
+This can also lead to a cascade of bugs where the root cause might not be obvious.
+
+For example: A method might concurrently create a list of numbers that is expected to be ordered but due to a concurrency bug the list contains unordered elements.
+This failure might only occur once in a thousand executions but if other methods depend on the correct order of elements in this list, the program might crash and the reason can be very hard to find.
+
+\subsection{Deadlocks}
+\begin{lstlisting}[float=h, language=Go, label=lst:deadlockWG, caption=Deadlock caused by waiting for the \emph{WaitGroup} at a wrong location -- based on \cite{tu2019go}]
+func main() {
+    var group sync.WaitGroup
+    group.Add(len(data))
+    for _, d := range data {
+        go func(d string) {
+            fmt.Printf("Processing: %s\n", d)
+            defer group.Done()
+        }(d)
+        group.Wait() // Causing the bug
+    }
+}
+\end{lstlisting}
+
+\begin{lstlisting}[float, language=Go, label=lst:deadlockCh, caption={Deadlock caused by misuse of an \emph{unbuffered Channel}}]
+func main() {
+    ch := make(chan int)
+    pr := make(chan int)
+    ch <- 1 // Program already blocks here
+    go func() {
+        i := <-ch
+        pr <- i
+    }()
+    <-pr
+}
+\end{lstlisting}
+
+One example for blocking bugs are \emph{deadlocks} where circular dependencies between resources block the flow of a program.
+\Cref{lst:deadlockWG} shows one example of such a deadlock in a Go program.
+The problem is a \emph{blocking synchronization} where the \lstinline{group.Wait()} inside the for-loop is causing the block.
+
+A second example of a deadlock that might not be obvious is \Cref{lst:deadlockCh} which uses two unbuffered channels to transfer information between threads.
+The problem here is that without an active listener on an unbuffered channel any send action will be blocked.
+To fix this one should use a buffered channel so the execution flow of the program can continue.
+
+% TODO: Put this somewhere else
+Another common problem in event-driven concurrent programs are \emph{blocking operations} like filesystem operations that are executed inside an event-handler.
+These ``can penalize and even paralyze the entire program execution.''~\cite{tchamgoue2012testing}
+
+\subsection{Data Races}
+\begin{lstlisting}[float=h, language=Go, label=lst:race, caption=Data race -- based on \cite{goRaceDetector}]
+func main() {
+	c := make(chan bool)
+	m := make(map[string]string)
+	go func() {
+		m["1"] = "a" // First conflicting access.
+		c <- true
+	}()
+	m["2"] = "b" // Second conflicting access.
+	<-c
+}
+\end{lstlisting}
+
+Data races are part of the non-blocking concurrency bugs.
+They occur when multiple threads simultaneously try to access a shared variable where at least one access is a write operation.~\cite{serebry2009threadsanitizer}
+
+\Cref{lst:race} shows an example of a data race that is frequently found.~\cite{serebry2009threadsanitizer}
+The data race happens when ``two threads access a non-thread-safe complex object [e.g. a map] without synchronization.''~\cite{serebry2009threadsanitizer}
+Even though the two threads in this example write to different keys, this might cause a corruption of data or even crash the program.
+
+\subsection{Atomicity and Order Violations}
+\begin{lstlisting}[float=h, language=Go, label=lst:order, caption=Test-and-Use bug pattern -- Order violation]
+func main() {
+    data := getData()
+    go func() {
+        process(data)
+        data = Nil
+    }()
+    if (data != Nil) {
+        process(data) // Might already be Nil
+    }
+}
+\end{lstlisting}
+
+Atomicity and order violations are concurrency bugs where the interleaving of threads violate the programmer's intention of atomicity and order.
+This often happens because:
+``Most programmers think sequentially and therefore they make mistakes easily when writing concurrent programs.''~\cite{lu2008mistakes}
+
+\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''.
+The programmer's intention is to check if a variable is not Nil and then use this variable.
+However, due to the thread that was launched before, it could happen that after the check in line 7, the thread of the goroutine gets scheduled and the data variable is set to Nil.
+
+\begin{lstlisting}[float=h, language=Go, label=lst:atomicity, caption=Load-Store bug pattern -- Atomicity violation]
+func main() {
+    var group sync.WaitGroup
+    group.Add(10)
+    sum := 0
+    for i = 0; i < 10; i++ {
+        go func() {
+            defer group.Done()
+            sum++ // Not atomic
+        }()
+    }
+    group.Wait()
+}
+\end{lstlisting}
+
+\Cref{lst:atomicity} shows one example of the infamous ``Load-Store'' bug pattern.
+The programmer assumes that ++ is an atomic operation because it is one literal in Go.
+However, after compilation this is expanded to 3 instructions: LOAD, INCREMENT and finally STORE.
+The thread scheduler could switch the context after any of these instructions what leads to undefined behavior.
+
+Lu, Park, Seo and Zhou conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs.~\cite{lu2008mistakes}
+One key finding was that:
+``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}
+
+% ------------------------------------ %
+% ------ DYNAMIC CODE ANALYSIS ------- %
+% ------------------------------------ %
+\section{Dynamic Code Analysis}
+\label{sct:dynamic}
+
+One of the main techniques for finding concurrency bugs is dynamic code analysis.
+In dynamic code analysis the runtime behavior of a program is observed to detect problematic patterns.
+
+One popular implementation of this is ``Record and Replay'' where the path of execution of an application is recorded and can later be deterministically replayed.
+This way programmers can easily find and fix those bug because, ``detecting concurrency bugs is difficult but once detected; correcting them is somehow an easy job.''~\cite{tchamgoue2012testing}
+To create a recording that contains all information to replay a program exactly the same way it was executed, the recorder needs to keep track of the schedule of the program as well as all variables that might not be deterministically reproducible.
+Like expected, this creates a lot of computational overhead in CPU time as well as storage.
+
+To track the schedule of a natively compiled program such as Go programs, the record tool has to sit between the host operating system and the application.
+One requirement to use such a tool in production is the ease of deployability which means that the operational overhead should be very small.
+One example for a tool that fulfills this requirement is \emph{rr}~\footnote{\url{https://rr-project.org/}} developed by \emph{Mozilla}.
+Their approach is to run the program on a fixed CPU core and track the system calls of the program by using \emph{\_ptrace\_}.
+This way they can record programs on an unmodified Linux system and without compiling instrumentation into the programs code which might affect the concurrency behavior.
+The main performance bottleneck are the context switches induced by ptrace.
+
+Record and replay could therefore be a great tool to monitor applications in production.
+Once a bug occurs and is either reported by a user or by the program itself, the programmer could just replay the thread schedule to reproduce, find and therefore fix the bug quickly.
+Additionally, this technique can help to reproduce all kind of concurrency bugs no matter if they are blocking or non-blocking.
+For programs that do not heavily depend on parallelism this approach is also very feasible.
+The tool \emph{rr} can record such programs with a performance overhead below the factor of two.
+However, for programs that heavily rely on parallelism the overhead to record and store the thread schedule can decrease the performance by more than a factor of 30.~\cite{o2017engineering}
+Especially for multi-threaded server applications that expect a high workload and where speed is crucial for the business, this approach is not feasible in production, yet.
+And even if the workload is low but the application is time-sensitive for example because of protocol timeouts, this technique is not applicable.
+
+Another set of tools for dynamic code analysis are dynamic data race detectors.
+As the name implies, these tools detect data races but because atomicity and order violations are often correlated, these tools can detect a variety of non-blocking concurrency bugs.
+There are multiple approaches for data race detection but in this case we refer to on-the-fly and post-mortem data race detection which are often referred to as dynamic.~\cite{serebry2009threadsanitizer}
+One data race detector that is mainly used for Go programs is the \emph{ThreadSanatizer} developed by Google.
+% TODO: more about thread sanatizer performance
+
+But ``data-race free does not mean concurrency bug free.''~\cite{lu2008mistakes}
+
+% ------------------------------------ %
+% ---- CONCURRENCY-AWARE TESTING ----- %
+% ------------------------------------ %
+\section{Concurrency-aware Testing}
+\label{sct:testing}
+
+Extensive testing has proven to minimize the bugs of software in production.\cite{makinen2014testing}
+However, traditional testing mostly covers sequential errors and can not detect concurrency bugs effectively.\cite{lu2008mistakes}
+This lies also in the nature of these bugs.
+Only certain thread interleavings trigger the bug and therefore the tests could complete successfully almost every time without ever triggering the bug.
+So new concurrency-aware tests are needed to find those bugs reliably.
+One method of concurrency aware testing has been proposed by Choi and Zeller.\cite{acm2002}
+% TODO: Reference for first occurrence of record and replay needed
+They make use of a technique called ``Record and Replay'' that will be covered more in section \ref{sct:dynamic}
+
+% ------------------------------------ %
+% ------ STATIC CODE ANALYSIS -------- %
+% ------------------------------------ %
+\section{Static Code Analysis}
+\label{sct:static}
+
+Finally, static code analysis
+
+% ------------------------------------ %
+% ------------ CONCLUSION ------------ %
+% ------------------------------------ %
+\section{Conclusion}
+\label{sct:conclusion}
+
+Dynamic code analysis has the main disadvantage that it can cost a lot of overhead when monitoring is active in production.
+Especially programs that heavily rely on multi-threaded execution like web-servers  can become very inefficient.
+
+Static code analysis has the advantage that it can run independently from production servers for example inside a CI.
+However, a high accuracy in detecting concurrency bugs requires a lot of computational resources due to the exponential growth of possible thread interleavings.
+
+Concurrency-aware testing
 

### 2020-05-10 src/paper.tex
Add more content to the main paper

--- 2020-04-07 src/paper.tex
@@
 \begin{abstract}
+% TODO: Why in production?
+Today's computer architectures rely more and more on parallel computing.
+Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
+To utilize the full performance of the given hardware, programmers need to write their software with a high degree of parallelism.
+This introduces the potential for bugs that would not occur in sequential programs due to non-determinism in the order of execution.
+For this purpose tools and languages have been developed to ease the creation of concurrency-aware programs and help programmers find concurrency bugs effectively.
+This paper will evaluate the different tools and techniques that are available to find, reproduce and fix concurrency bugs from the viewpoint of how usable they are in a production environment.
+The requirements for these tools are that they need to be easy to deploy, they should bring only a minimal computational and storage overhead, they should have a high coverage with minimal false-positive reports and they need to enable the developer to quickly find the cause of a bug.
+All tools are evaluated for the Go programming language but the concepts are mainly applicable to every other language as well.
 \end{abstract}
@@
 \begin{IEEEkeywords}
+    [Software Engineering]: Testing and Debugging
 \end{IEEEkeywords}
 
+% ------------------------------------ %
+% ----------- INTRODUCTION ----------- %
+% ------------------------------------ %
 \section{Introduction}
-Test \cite{acm2002}
+
+\subsection{Heisenbugs}
+
+\begin{quote}
+``Everyone knows that debugging is twice as hard as writing a program in the first place.
+So if you're as clever as you can be when you write it, how will you ever debug it?'' --- Brian W.
+Kernighan\cite{kernighan1974elements}
+\end{quote}
+
+Writing concurrent programs can be very challenging.
+Computational work gets distributed among different processes and threads that need to synchronize to be able to speed up execution time on multi-core architectures.
+Due to this complexity it is no surprise that bugs that are correlated to concurrency and parallel program execution occur quite often.
+These bugs are also the most difficult to debug because they are so complex in their nature and not easy to reproduce.~\cite{tu2019go}
+This is because of the non-determinism that is introduced when multiple threads or processes are executed simultaneously.
+In contrast to sequential programs that only run on one thread and process, concurrent programs have multiple paths of execution that are executed more or less independently from each other.
+The order these threads are executed is determined by the operating system's scheduler.
+Because a developer, like the program itself, cannot predict what resources also need processing time on a machine, the concrete schedule has to be handled as non-predictable and therefore as non-deterministic.
+But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread executions or thread interleavings appear, cannot be reliably reproduced without knowing the schedule it appeared in.
+
+In 1986 Jim Gray introduced the term ``Heisenbug'' for these kind of software failures.
+He wrote:
+
+\begin{quote}
+``The assertion that most production software bugs are soft -- Heisenbugs that go away when you look at them -- is well known to systems programmers.
+Bohrbugs, like the Bohr atom, are solid, easily detected by standard techniques, and hence boring.
+But Heisenbugs may elude a bugcatcher for years of execution.
+Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.
+This is analogous to the Heisenberg Uncertainty Principle in Physics.''\cite{gray1986computers}
+\end{quote}
+
+The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix or even avoid them.~\cite{tu2019go}
+Thus, numerous tools and techniques have been developed that propose different solutions to solve the problem of concurrency bugs and accelerate the debugging process.
+
+This paper will evaluate some of these techniques and tools and how they can be used to ease the process of debugging concurrent bugs.
+The main focus is to detect bugs efficiently in production, which means that developers should be able to fix bugs with a minimum amount of work and time.
+Also, the manifestation of concurrency bugs should be reported immediately with minimal overhead of computational resources.
+For all of this it is also desirable to have a maximum amount of coverage with only minimal false-positive reports.
+
+\subsection{The Go Programming Language}
+All code examples are written in Go which is a statically typed programming language developed by Google, which compiles down to single native binaries.
+The reason we are using Go and not C/C++ or Java is the recent popularity of it for writing highly concurrent programs.
+Popular open-source projects like \emph{Kubernetes}~\footnote{\url{https://kubernetes.io/}}, \emph{Docker}~\footnote{\url{https://www.docker.com/}} or \emph{Prometheus}~\footnote{\url{https://prometheus.io/}} are all written in Go.
+In the official documentation of Go it claims: ``Go is expressive, concise, clean, and efficient.
+Its concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines, while its novel type system enables flexible and modular program construction.''~\cite{goDocs}
+So, Go tries to solve some of the problems that arise when writing concurrent software application by their design of language.
+% TODO: How does Go want to archive this?
+Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} followed by a named or anonymous function.
+The Go runtime maps these goroutines to normal Kernel threads during execution.
+Go encourages programmers to use concurrency by \emph{message passing}, which is assumed to be safer and more convenient to use than \emph{shared memory}.
+However, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
+A study by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
+They came to the conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
+Projects written in Go also tend to have more concurrency than projects written in traditional programming languages such as C.~\cite{tu2019go}
+Given all those conclusions it is even more important for Go developers to catch concurrency bugs effectively.
+Together with the increasing popularity of the language, the demand for production ready tools to catch concurrency bugs is very high, which is why this paper evaluates existing techniques in the context of the Go programming language.
+
+\subsection{Table of Contents}
+
+\Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes.
+\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs and reliably reproduce them.
+\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically by manipulating the thread scheduler.
+And \Cref{sct:static} finally covers how to detect concurrency bugs with static code analysis.
+In the end there is a conclusion with a short outlook on the possible future of multi-threaded debugging.
+
+% ------------------------------------ %
+% --- TAXONOMY OF CONCURRENCY BUGS --- %
+% ------------------------------------ %
+\section{Taxonomy of Concurrency Bugs}
+\label{sct:taxonomy}
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figures/ConcurrencyBugClasses.pdf}
+    \caption{A Taxonomy of Concurrency Bugs -- based on\cite{tchamgoue2012testing}}
+    \label{fig:classes}
+\end{figure}
+
+\Cref{fig:classes} shows a taxonomy of concurrency bugs based on the work of Tchamgoue, Kim and Jun.~\cite{tchamgoue2012testing}
+They distinguish between two main categories of concurrency bugs: \emph{Blocking} and \emph{non-blocking}.
+Blocking bugs are bugs where the execution of a program is unintentionally blocked and the program cannot terminate.
+% TODO: Meh, don't know if that is right..
+The manifestation of these kind of bugs is very noticeable because the whole application freezes and has to be shutdown before the computation can continue.
+Non-blocking bugs are often harder to find because they can occur even when the termination of a program was successful but the result is wrong.
+This can also lead to a cascade of bugs where the root cause is a non-blocking concurrency bug which might not be obvious.
+
+For example: A method concurrently creates a list of numbers that is expected to be ordered but due to a non-blocking concurrency bug the list contains unordered elements.
+This failure might only occur once in a thousand executions due to the exponentially growing number of possible thread interleavings.
+But if other methods depend on the correct order of elements in this list, the program might crash or generate wrong results and the reason can be very hard to find.
+
+\subsection{Deadlocks}
+\begin{lstlisting}[float=h, language=Go, label=lst:deadlockWG, caption=Deadlock caused by waiting for the \emph{WaitGroup} at a wrong location -- based on \cite{tu2019go}]
+func main() {
+    var group sync.WaitGroup
+    group.Add(len(data))
+    for _, d := range data {
+        go func(d string) {
+            fmt.Printf("Processing: %s\n", d)
+            defer group.Done()
+        }(d)
+        group.Wait() // Causing the bug
+    }
+}
+\end{lstlisting}
+
+The most commonly manifestation of blocking bugs are \emph{deadlocks}, where circular dependencies between resources block the flow of a program.
+\Cref{lst:deadlockWG} shows one example of such a deadlock in a Go program.
+The problem is a \emph{blocking synchronization} where the \lstinline{group.Wait()} inside the for-loop is causing the block.
+Although the error seems obvious in this case, those small mistakes can quickly happen and can get unrecognized into the production environment if not tested well enough.
+
+\begin{lstlisting}[float=h, language=Go, label=lst:deadlockCh, caption={Deadlock caused by misuse of an \emph{unbuffered Channel}}]
+func main() {
+    ch := make(chan int)
+    pr := make(chan int)
+    ch <- 1 // Program already blocks here
+    go func() {
+        i := <-ch
+        pr <- i
+    }()
+    <-pr
+}
+\end{lstlisting}
+
+A second example of a deadlock that might not be obvious is \Cref{lst:deadlockCh} which uses two unbuffered channels to transfer information between threads.
+The problem here is that without an active listener on an unbuffered channel, any send action will be blocked.
+To fix this, one could replace the unbuffered channel with a buffered one so that the execution flow of the program can continue without blocking.
+
+Another common problem in event-driven concurrent programs are \emph{blocking operations} like filesystem operations that are executed inside an event-handler.
+These ``can penalize and even paralyze the entire program execution.''~\cite{tchamgoue2012testing}
+
+\subsection{Data Races}
+\begin{lstlisting}[float=h, language=Go, label=lst:race, caption=Data race -- based on \cite{goRaceDetector}]
+func main() {
+	c := make(chan bool)
+	m := make(map[string]string)
+	go func() {
+		m["1"] = "a" // First conflicting access.
+		c <- true
+	}()
+	m["2"] = "b" // Second conflicting access.
+	<-c
+}
+\end{lstlisting}
+
+Data races are part of the non-blocking concurrency bugs.
+They occur when multiple threads simultaneously try to access a shared variable where at least one access is a write operation.~\cite{serebry2009threadsanitizer}
+
+\Cref{lst:race} shows an example of a data race that is frequently found.~\cite{serebry2009threadsanitizer}
+The data race happens when ``two threads access a non-thread-safe complex object [e.g. a map] without synchronization.''~\cite{serebry2009threadsanitizer}
+Even though the two threads in this example write to different keys, this might cause a corruption of data or even crash the program because the default Go map is not concurrency-aware.
+
+% TODO: Is it a data race or an atomicity violation?
+A special case of data races are multi-variable data races.
+In this case multiple variables that are semantically correlated are accessed by multiple threads and could therefore loose their semantic correctness when not handled correctly.
+A common example for this are structs that have a data and a length field.
+Whenever the data field gets updated, so does the length field which is semantically correlated with the data.
+When multiple threads update these variables they need to synchronize not only on each of the variables separately but on all semantically correlated variables at the same time.~\cite{lu2007muvi}
+
+\subsection{Atomicity and Order Violations}
+\begin{lstlisting}[float=h, language=Go, label=lst:order, caption=Test-and-Use bug pattern -- Order violation]
+func main() {
+    data := getData()
+    go func() {
+        process(data)
+        data = nil
+    }()
+    if (data != nil) {
+        process(data) // Might already be nil
+    }
+}
+\end{lstlisting}
+
+Atomicity and order violations are concurrency bugs where the interleaving of threads violate the programmer's intention of atomicity and order.
+This often happens because:
+``Most programmers think sequentially and therefore they make mistakes easily when writing concurrent programs.''~\cite{lu2008mistakes}
+
+\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''.
+The programmer's intention is to check if a variable is not \lstinline{nil} and then use this variable.
+However, due to the thread that was launched before, it could happen that after the check in line 7, the thread of the goroutine gets scheduled and the data variable is set to \lstinline{nil}.
+
+\begin{lstlisting}[float=h, language=Go, label=lst:atomicity, caption=Load-Store bug pattern -- Atomicity violation]
+func main() {
+    var group sync.WaitGroup
+    group.Add(10)
+    sum := 0
+    for i = 0; i < 10; i++ {
+        go func() {
+            defer group.Done()
+            sum++ // Not atomic
+        }()
+    }
+    group.Wait()
+}
+\end{lstlisting}
+
+\Cref{lst:atomicity} shows one example of the infamous ``Load-Store'' bug pattern.
+The programmer assumes that ++ is an atomic operation because it is one literal in Go.
+However, after compilation this is expanded to 3 instructions: LOAD, INCREMENT and finally STORE.
+The thread scheduler could switch the context after any of these instructions what leads to undefined behavior, when multiple threads try to increment the same variable.
+
+Lu, Park, Seo and Zhou conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs.~\cite{lu2008mistakes}
+One key finding was that:
+``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}
+
+% ------------------------------------ %
+% ------ DYNAMIC CODE ANALYSIS ------- %
+% ------------------------------------ %
+\section{Dynamic Code Analysis}
+\label{sct:dynamic}
+
+One of the main techniques for finding concurrency bugs is dynamic code analysis.
+% TODO: Find better word for problematic
+In dynamic code analysis the runtime behavior of a program is observed to detect problematic patterns like unsynchronized variable accesses or deadlocks.
+
+One popular implementation of this is ``Record and Replay'' where the path of execution of an application is recorded and can later be deterministically replayed.
+This way developers can easily find and fix those bug because, ``detecting concurrency bugs is difficult but once detected; correcting them is somehow an easy job.''~\cite{tchamgoue2012testing}
+To create a recording that contains all information to replay a program exactly the same way it was executed, the recorder needs to keep track of the schedule of the program as well as all variables that might not be deterministically reproducible.
+Furthermore, any source of non-determinism such as the the interaction with the network or graphics card or any call to external libraries need to be recorded.~\cite{lidbury2019sparse}
+Like expected, this creates a lot of computational overhead in CPU time as well as storage.
+This approach also forms the foundation for many other techniques and has many different variants and implementations.
+
+To track the schedule of a natively compiled program such as Go programs, the record tool has to sit between the host operating system and the application.
+One requirement to use such a tool in production is the ease of deployability which means that the operational overhead should be very small.
+There have been approaches with custom hardware or patched operating system kernels to hijack the scheduler, but these setups are quite complex and not easy to deploy.
+One example for a tool that fulfills this requirement is \emph{rr}~\footnote{\url{https://rr-project.org/}} developed by \emph{Mozilla}.
+Their approach is to run the program to observe on a fixed CPU core and track the system calls by using \emph{\_ptrace\_}.
+This way they can record programs on an unmodified Linux system and without compiling instrumentation into the programs code which might affect the concurrency behavior.
+The main performance bottleneck beside the execution on only one CPU core, are the context switches induced by ptrace.~\cite{o2017engineering}
+Lidbury and Donaldson~\cite{lidbury2019sparse} examined further ways to reduce the computational overhead by a sparse recording approach.
+By selectively ignoring ``unimportant'' sources of non-determinism they were able to record input and output heavy applications like video games with an acceptable overhead where as \emph{rr} was not able to record at all due to the massive overhead of recording every source of non-determinism.
+During replay they execute the thread schedule as recorded and mock every source of recorded non-determinism but let the ignored regions of the program execute non-deterministically.
+
+Record and replay could therefore be a great tool to monitor applications in production.
+Once a bug occurs and is either reported by a user or by the program itself, the programmer could just replay the thread schedule to reproduce, find and therefore fix the bug quickly.
+Additionally, this technique can help to reproduce all kind of concurrency bugs, no matter if they are blocking or non-blocking.
+For programs that do not heavily depend on parallelism this approach is also very feasible.
+The tool \emph{rr} can record such programs with a performance overhead below the factor of two.
+However, for programs that heavily rely on parallelism the overhead to record and store the thread schedule and all necessary parameters can decrease the performance by more than a factor of 30.~\cite{o2017engineering}
+Especially for multi-threaded server applications that expect a high workload and where speed is crucial for the business, this approach is not feasible in production, yet.
+And even if the workload is low but the application is time-sensitive for example because of protocol timeouts, it is not applicable without extending the acceptable timeouts.
+
+Another set of tools for dynamic code analysis are dynamic data race detectors.
+As the name implies, these tools detect data races but because atomicity and order violations are often correlated, these tools can detect a variety of non-blocking concurrency bugs.
+There are multiple approaches for data race detection but in this case we refer to on-the-fly and post-mortem data race detection which are often referred to as dynamic.~\cite{serebry2009threadsanitizer}
+One data race detector that is mainly used for Go programs is the \emph{ThreadSanitizer} developed by Google.
+It is built in the Golang compiler tool-chain and can be toggled with a simple compiler flag which makes it very easy to use.
+``The tool performs compile-time instrumentation of the source program, in which all (atomic and non-atomic) accesses to potentially shared locations, as well as fence operations, are instrumented with calls into a statically linked run-time library.''~\cite{lidbury2019sparse}
+To distinguish between synchronized accesses and data races during runtime, the compiler also has to inject dynamic annotations into the program.
+With these annotations the data race detector understands the synchronization primitives used by the programming language.
+When the program along with the race detector gets executed, the data race detector ``observes the program execution as a sequence of events.''~\cite{serebry2009threadsanitizer}
+To detect a data race, there are two algorithms that are mainly used among all detectors: \emph{happens-before} and \emph{lockset}.
+Locking events are Write and Read Locks / Unlocks.
+Happens-before events are Signal and Wait.
+The \emph{ThreadSanitizer} uses an hybrid approach by combining both algorithms into one and observing events of both categories.
+
+Performance-wise the race detection is unfortunately not better than the record and replay approach.
+The official race detector article on the Go website claims: ``The cost of race detection varies by program, but for a typical program, memory usage may increase by 5-10x and execution time by 2-20x.''~\cite{goRaceDetector}
+Which is in fact better than the original 20x-50x evaluated by Serebryany and Iskhodzhanov in 2009.~\cite{serebry2009threadsanitizer}
+However, a possible improvement could be to ignore certain uncritical regions and just observe hotspots where data races are suspected.
+
+Beside their ease of use, one of the main disadvantages of dynamic analysis tools is their coverage, that `` [...] is typically small since only a few executions are explored.''~\cite{qadeer2004kiss}
+Additionally, they depend on the manifestation and especially the recognition of these bugs to report them correctly.
+
+% ------------------------------------ %
+% ---- CONCURRENCY-AWARE TESTING ----- %
+% ------------------------------------ %
+\section{Concurrency-aware Testing}
+\label{sct:testing}
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{figures/Concurrency-Testing.pdf}
+    \caption{Automatic concurrency-aware testing for Go programs -- based on\cite{acm2002}}
+    \label{fig:testing}
+\end{figure}
+
+Extensive testing has proven to minimize the bugs of software in production.\cite{makinen2014testing}
+However, traditional testing mostly covers sequential errors and cannot detect concurrency bugs effectively.\cite{lu2008mistakes}
+This lies in the nature of these bugs themself.
+When only certain thread interleavings trigger the bug, the tests could complete successfully almost every time without ever finding a manifestation.
+So, new concurrency-aware tests are needed to find concurrency bugs reliably.
+
+One early attempt to test programs for concurrency bugs was to inject delays at random points in the program code, but it is yet unclear if this really gives an acceptable bug coverage.~\cite{lu2008mistakes}
+Furthermore, the tests should try to reduce the perturbation on the execution, so the software can be evaluated the same way as it would run in a regular deployment in production.
+
+Another method of concurrency-aware testing has been proposed by Choi and Zeller.\cite{acm2002}
+They implemented a tool called ``DEJAVU'' for the \emph{Jalapeño} Java Virtual Machine but the concept is also applicable to tools that work with Go or any other programming language.
+They make use of a technique called \emph{Delta debugging}, which describes the process of narrowing down the spot where a failure is introduced by going back and forth between working and non-working conditions, in this case working and non-working thread schedules.~\cite{zeller2002delta}
+\Cref{fig:testing} shows a modified model of the \emph{DEJAVU} approach, suited for Go programs by using rr instead of of \emph{DEJAVU} for the recording and replay of the thread schedule.
+As a prerequisite, test cases need to be created that define the expected behavior of the application.
+These can either succeed or fail which can be shown be the exit code for example.
+In the first step, the Go application with these test cases gets executed and the thread schedule gets recorded by \emph{rr}.
+The either failing or succeeding recorded thread schedule is then passed to the next instance where the delta debugging unit requests new executions of the application with an alternated thread schedule until it can pinpoint the failure-inducing thread switch.
+The thread schedule is generated by the \emph{Schedule generator} and is replayed by the replay module in \emph{rr}.
+The schedule as well as the test result is passed back to the delta debugging unit.
+``By systematically narrowing down the difference between a thread schedule that makes the program pass and another schedule that makes the program fail, the Delta Debugging approach can pinpoint the error location automatically -- namely the location(s) where a thread switch causes the program to fail.''~\cite{acm2002}
+
+While this technique cannot be used in a productive deployment setup, it is intended to extend existing tests that for example run inside a CI pipeline.
+One major drawback however, is the number of thread interleavings that can grow exponentially with the number of executed threads.
+But for most cases it is not necessary to evaluate all possible thread schedules.
+Lu et al.
+have found in their study of real-world concurrency bugs that ``Almost 96\% of the examined concurrency bugs are guaranteed to manifest if certain partial order between 2 threads is enforced.''~\cite{lu2008mistakes}
+Which means that ``Pairwise testing on concurrent program threads can expose most concurrency bugs, and greatly reduce the testing complexity.''~\cite{lu2008mistakes}
+To increase the coverage of concurrency bugs, this approach could also be combined with a data race detector to dynamically detect data races within certain replays of a thread schedule.
+This way the test cases can be minimized to semantical correctness checks, which decreases the operational overhead for such a setup.
+The setup of such a system also comes with a relatively low operational overhead because once such a system is build, it could be easily embedded within the CI pipelines of a project.
+
+% ------------------------------------ %
+% ------ STATIC CODE ANALYSIS -------- %
+% ------------------------------------ %
+\section{Static Code Analysis}
+\label{sct:static}
+
+Static code analysis describes the technique of predicting the behavior of a program by analyzing the source code without compiling nor executing it.
+
+A common approach to static code analysis is model-checking where the code gets translated into mathematical models such as state machines or graphs.
+But model checking for concurrent programs is complex due to the number of possible thread interleavings that can grow exponentially with the number of executed threads.
+Therefore, static data race detectors for example are unfeasible for very large code bases.~\cite{serebry2009threadsanitizer}
+
+One implementation of static code analysis to detect concurrency bugs is sequentialization as proposed by Qadeer and Wu.~\cite{qadeer2004kiss}
+Their approach is to transform a parallel program into a sequential one and simulate a large subset of the behavior of the parallel program on the sequential one.
+This way they can use traditional and highly optimized sequential model-checkers to analyze the application.
+By this approach the system has no false-positive reports on concurrency bugs but the overall coverage is not very high.
+
+Although unfeasible for detecting data races in large code bases, there are multiple implementations of static deadlock-detectors for concurrent programs.
+One concrete static deadlock detector for the Go programming language is called \emph{Godel Checker}.~\cite{godelChecker}
+It uses that Go's message passing concurrency mechanisms by channels is ``[...] greatly inspired by advances in formal languages for concurrency theory known as process calculi.''~\cite{lange2018verification}
+The \emph{Godel Checker} tool extracts the main concurrency characteristics of the program into a mathematical model called \emph{Behavioural Types}.
+These types can be further processed into a \emph{static single assignment}, which then can be processed by classical model and termination checker.
+This way not only total deadlocks where all threads are stuck, but also partial deadlocks and channel safety threats can be found.
+
+The computational overhead for this approach is also surprisingly low.
+A program with 18 kLOC can be checked within minutes.~\cite{lange2018verification}
+Unfortunately this technique has quite a few limitations.
+Only immutable Go channels are supported that are not located in a dynamic data type.
+Furthermore, Go does not force the developer to use synchronization by message passing, so any deadlocks introduced by traditional locking mechanisms will be not be noticed.
+
+% ------------------------------------ %
+% ------------ CONCLUSION ------------ %
+% ------------------------------------ %
+\section{Conclusion}
+\label{sct:conclusion}
+
+While having multi-threaded applications is convenient and also necessary in today's world of computation, the non-determinism makes it hard to find, reproduce and therefore fix concurrency bugs.
+Although there is a variety of tools that try to ease the process of multi-threaded debugging, as shown, none of them really fulfills all requirements for a production-ready drop-in solution.
+
+Dynamic code analysis has the main disadvantage that it can cost a lot of overhead when monitoring is active in a production environment.
+Especially programs that heavily rely on multi-threaded execution like web-servers  can become very inefficient or could even fail due to timeout limitations.
+Dynamic data race detectors could eventually be optimized to an acceptable overhead by ignoring most of the assumed-to-be-safe code regions, but this also drastically lowers the overall coverage.
+Beside that, ``data-race free does not mean concurrency bug free.''~\cite{lu2008mistakes}
+The main advantage of the dynamic approach is that it theoretically could be deployed to a real production environment.
+For programs that are mainly event-driven this would give the ability to observe the real-world behavior in a reproducible way.
+
+Concurrency-aware testing is very promising in the regard of extending an existing testing pipeline.
+This way concurrency bugs could be detected even before the software is deployed into production.
+The overhead can be minimized by optimizing the thread schedule generator and by ignoring most uncritical regions of the code.
+However, the coverage depends on the quality of the tests that also need to be aware of multi-variable semantic data races.
+
+Static code analysis is only feasible in small code bases, due to the exponentially growing thread interleavings that need to be explored, which requires a lot of computational overhead.
+Additionally multi-variable accesses of only semantically correlated variables need manual annotations or configuration.
+``Unfortunately, existing techniques cannot effectively extract such semantic correlations.
+Traditional compiler analysis cannot catch them, because many correlated variables are just semantically correlated and do not necessarily have data dependencies, [...]''~\cite{lu2007muvi}
+Beside that there are some promising tools for statically finding deadlocks by translating programs into a mathematical model.
+
+Concluding this paper, the evaluation of the different approaches shows that there is not one tool to solve all problems that arise when writing concurrent programs.
+But there are multiple techniques that can be combined to at least get some support in debugging multi-threaded applications.
+
+However, for modern micro-service architectures that get more and more popular, these tools could actually be useful because the code bases tend to get smaller.
+That means that the overhead of instrumenting, testing or model-checking these applications gets smaller in absolute numbers.
+Record and replay could also be deployed to just some replicas of an instance, reducing the impact on crucial business speed whilst collecting useful information to optimize the whole code base.
+
+But then again, detecting and reporting concurrency bugs is just the first step in fixing them.
+Developers need to understand the reporting and need to distinguish between real bugs and false-positives.
+So, ``There is no silver bullet for fixing concurrency bugs.''~\cite{lu2008mistakes}
 

### 2020-05-10 src/paper.tex
Add fixes for bug examples and STM

@@
 This introduces the potential for bugs that would not occur in sequential programs due to non-determinism in the order of execution.
-For this purpose tools and languages have been developed to ease the creation of concurrency-aware programs and help programmers find concurrency bugs effectively.
-This paper will evaluate the different tools and techniques that are available to find, reproduce and fix concurrency bugs from the viewpoint of how usable they are in a production environment.
+Such bugs can be very challenging to fix because they are not easily reproducible.
+For this purpose tools and languages have been developed to ease the creation of concurrency-aware programs and to help programmers find concurrency bugs effectively.
+This paper will evaluate three different techniques and concrete tool implementations of them, that are available to find, reproduce and fix concurrency bugs from the viewpoint of how usable they are in a production environment.
 The requirements for these tools are that they need to be easy to deploy, they should bring only a minimal computational and storage overhead, they should have a high coverage with minimal false-positive reports and they need to enable the developer to quickly find the cause of a bug.
+The techniques evaluated are: Dynamic code analysis with the record and replay tool rr and the data race detector ThreadSanitizer.
+Concurrency-aware testing with a combined approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
+Static code analysis with the methods of sequentialization and model-checking.
 All tools are evaluated for the Go programming language but the concepts are mainly applicable to every other language as well.
@@
 \begin{IEEEkeywords}
-    [Software Engineering]: Testing and Debugging
+    [Software Engineering]: Software testing and Debugging
+    [Computing methodologies]: Concurrent programming languages
 \end{IEEEkeywords}
@@
 
-\Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes.
-\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs and reliably reproduce them.
-\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically by manipulating the thread scheduler.
-And \Cref{sct:static} finally covers how to detect concurrency bugs with static code analysis.
-In the end there is a conclusion with a short outlook on the possible future of multi-threaded debugging.
+\Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes: Deadlocks, Data races and Atomicity and Order violations.
+\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs, reliably reproduce them by utilizing record and replay with the tool \emph{rr} and detect data races dynamically with the tool \emph{ThreadSanitizer}.
+\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically by manipulating the thread scheduler and delta debugging.
+And \Cref{sct:static} finally covers how to detect concurrency bugs with static code analysis by sequentialization and model-checking.
+In the end there is a conclusion with a short comparison of all techniques and a quick outlook on the possible future of multi-threaded debugging.
 
@@
 
-For example: A method concurrently creates a list of numbers that is expected to be ordered but due to a non-blocking concurrency bug the list contains unordered elements.
-This failure might only occur once in a thousand executions due to the exponentially growing number of possible thread interleavings.
-But if other methods depend on the correct order of elements in this list, the program might crash or generate wrong results and the reason can be very hard to find.
+For example: A method concurrently creates a list of numbers that is expected to be ordered but due to a non-blocking concurrency bug the list suddenly contains unordered elements.
+This failure might only occur once in a thousand executions or even less, due to the exponentially growing number of possible thread interleavings.
+But if other methods depend on the correct order of the elements in this list, the program might crash or generate wrong results and the reason can be very hard to find.
 
@@
 The problem is a \emph{blocking synchronization} where the \lstinline{group.Wait()} inside the for-loop is causing the block.
+This statement has to be moved outside the for loop to resolve the unintentional blocking and fix the concurrency bug.
 Although the error seems obvious in this case, those small mistakes can quickly happen and can get unrecognized into the production environment if not tested well enough.
@@
 To fix this, one could replace the unbuffered channel with a buffered one so that the execution flow of the program can continue without blocking.
+This shows how important it is to know the concrete implementation of a concurrency abstraction.
+Even though intended to ease the synchronization between threads and make inter-thread communication safer, by not knowing the implementation of an abstraction the developer can unknowingly create hard to find concurrency bugs.
 
@@
 The data race happens when ``two threads access a non-thread-safe complex object [e.g. a map] without synchronization.''~\cite{serebry2009threadsanitizer}
-Even though the two threads in this example write to different keys, this might cause a corruption of data or even crash the program because the default Go map is not concurrency-aware.
+Even though the two threads in this example write to different keys of the map \lstinline{m}, this might cause a corruption of data or even crash the program because the default Go map implementation is not concurrency-aware.
+To fix this, the access to the map needs to be synchronized by a lock for example.
 
@@
 The programmer's intention is to check if a variable is not \lstinline{nil} and then use this variable.
-However, due to the thread that was launched before, it could happen that after the check in line 7, the thread of the goroutine gets scheduled and the data variable is set to \lstinline{nil}.
+However, due to the thread that was launched before, it could happen that after the \lstinline{if} check in line 7, the thread of the goroutine gets scheduled and the data variable is set to \lstinline{nil}.
+To fix this bug, the check and the usage of the variable need to become an atomic operation to enforce the order of execution.
 
@@
 The thread scheduler could switch the context after any of these instructions what leads to undefined behavior, when multiple threads try to increment the same variable.
+To fix this bug pattern, the \lstinline{++} operation also needs to be replaced by an atomic operation that does not allow other threads to access the \lstinline{sum} variable while incrementing.
 
@@
 ``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}
+
+A promising solution to atomicity and order violation bugs is \emph{software transactional memory} (STM) as proposed by Peyton Jones.~\cite{peytonjones2007beautiful}
+It is an alternative to traditional lock-based synchronization where atomic regions get declared explicitly.
+Languages like Haskell provide STM by their design of language but Go can also utilize this mechanism by using external libraries.
 

### 2020-05-10 src/paper.tex
Small fixes for the first draft deadline

@@
 % TODO: How does Go want to archive this?
-Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} followed by a named or anonymous function.
+Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} keyword followed by a named or anonymous function.
 The Go runtime maps these goroutines to normal Kernel threads during execution.
@@
 However, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
-A study by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
+A study conducted by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
 They came to the conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
@@
 The problem here is that without an active listener on an unbuffered channel, any send action will be blocked.
-To fix this, one could replace the unbuffered channel with a buffered one so that the execution flow of the program can continue without blocking.
+To fix this, one could replace the unbuffered channel with a buffered one, so that the execution flow of the program can continue without blocking.
 This shows how important it is to know the concrete implementation of a concurrency abstraction.
@@
 As a prerequisite, test cases need to be created that define the expected behavior of the application.
-These can either succeed or fail which can be shown be the exit code for example.
+These can either succeed or fail which can be shown by the exit code for example.
 In the first step, the Go application with these test cases gets executed and the thread schedule gets recorded by \emph{rr}.

### 2020-05-30 src/paper.tex
Add new section and shorten the introduction

@@
 
-% TODO: Correctly align the caption with the listing frame
-\usepackage{caption}
-\DeclareCaptionFont{white}{\color{white}}
-\DeclareCaptionFormat{listing}{\colorbox{darkgray}{\parbox{\dimexpr\linewidth-0.7em\relax}{#1#2#3}}}
-\captionsetup[lstlisting]{format=listing,labelfont=white,textfont=white}
-
 \definecolor{codegreen}{rgb}{0,0.6,0}
@@
     breaklines=true,
-    captionpos=t,
+    captionpos=b,
     keepspaces=true,
@@
 \lstset{style=mystyle}
-\renewcommand{\lstlistingname}{Code}
-\Crefname{listing}{Code}{Codes}
-\crefname{listing}{code}{codes}
 
@@
 
+% ------------------------------------ %
+% ------------- ABSTRACT ------------- %
+% ------------------------------------ %
 \begin{abstract}
-% TODO: Why in production?
-Today's computer architectures rely more and more on parallel computing.
-Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
-To utilize the full performance of the given hardware, programmers need to write their software with a high degree of parallelism.
-This introduces the potential for bugs that would not occur in sequential programs due to non-determinism in the order of execution.
-Such bugs can be very challenging to fix because they are not easily reproducible.
-For this purpose tools and languages have been developed to ease the creation of concurrency-aware programs and to help programmers find concurrency bugs effectively.
-This paper will evaluate three different techniques and concrete tool implementations of them, that are available to find, reproduce and fix concurrency bugs from the viewpoint of how usable they are in a production environment.
-The requirements for these tools are that they need to be easy to deploy, they should bring only a minimal computational and storage overhead, they should have a high coverage with minimal false-positive reports and they need to enable the developer to quickly find the cause of a bug.
-The techniques evaluated are: Dynamic code analysis with the record and replay tool rr and the data race detector ThreadSanitizer.
-Concurrency-aware testing with a combined approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
-Static code analysis with the methods of sequentialization and model-checking.
-All tools are evaluated for the Go programming language but the concepts are mainly applicable to every other language as well.
+Concurrency bugs are often hard to find and therefore hard to fix.
+In this paper we evaluate different techniques and tools that are currently available to effectively find, fix and avoid concurrency bugs.
+The requirements for these tools are that they need to be easy to use, they should bring only a minimal computational and storage overhead and they should have a high coverage with minimal false-positive reports.
+In short: These tools should be production-ready and applicable to real-world software projects.
+The techniques evaluated are:
+\begin{itemize}
+    \item Dynamic code analysis with record and replay and dynamic data race detection.
+    \item Concurrency-aware testing with a combined approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
+    \item Static code analysis with the methods of sequentialization and model-checking.
+    \item Language and software design to avoid concurrency bugs
+\end{itemize}
 \end{abstract}
@@
 
-Writing concurrent programs can be very challenging.
-Computational work gets distributed among different processes and threads that need to synchronize to be able to speed up execution time on multi-core architectures.
-Due to this complexity it is no surprise that bugs that are correlated to concurrency and parallel program execution occur quite often.
-These bugs are also the most difficult to debug because they are so complex in their nature and not easy to reproduce.~\cite{tu2019go}
-This is because of the non-determinism that is introduced when multiple threads or processes are executed simultaneously.
-In contrast to sequential programs that only run on one thread and process, concurrent programs have multiple paths of execution that are executed more or less independently from each other.
-The order these threads are executed is determined by the operating system's scheduler.
+% Motivation
+Today's computer architectures rely more and more on parallel computing.
+Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
+To utilize the full performance of the given hardware, programmers need to write their software with a high degree of parallelism.
+In contrast to sequential programs which only have one deterministic path through the program code, concurrent programs have multiple paths of execution that are executed independently from each other.
+The order in which these threads are executed is determined by the operating system's scheduler.
+The resulting order of thread executions is often referred to as \emph{thread interleavings}.
 Because a developer, like the program itself, cannot predict what resources also need processing time on a machine, the concrete schedule has to be handled as non-predictable and therefore as non-deterministic.
-But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread executions or thread interleavings appear, cannot be reliably reproduced without knowing the schedule it appeared in.
+This introduces the potential for bugs that would not occur in sequential programs due to non-determinism in the order of execution.
+Such bugs can be very challenging to debug because they are not easily reproducible~\cite{tu2019go}.
+Due to this complexity, it is no surprise that bugs that are related to concurrency and parallel program execution occur quite often.
+But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread interleavings appear, cannot be reliably reproduced without knowing the schedule it appeared in.
 
@@
 \begin{quote}
-``The assertion that most production software bugs are soft -- Heisenbugs that go away when you look at them -- is well known to systems programmers.
-Bohrbugs, like the Bohr atom, are solid, easily detected by standard techniques, and hence boring.
-But Heisenbugs may elude a bugcatcher for years of execution.
-Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.
-This is analogous to the Heisenberg Uncertainty Principle in Physics.''\cite{gray1986computers}
+``But Heisenbugs may elude a bugcatcher for years of execution. Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.''\cite{gray1986computers}
 \end{quote}
 
-The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix or even avoid them.~\cite{tu2019go}
+Heisenbugs are bugs that only occur with very specific thread interleavings that might not occur when a bugcatcher evaluates the program.
+The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix and avoid them~\cite{tu2019go}.
 Thus, numerous tools and techniques have been developed that propose different solutions to solve the problem of concurrency bugs and accelerate the debugging process.
@@
 \subsection{The Go Programming Language}
-All code examples are written in Go which is a statically typed programming language developed by Google, which compiles down to single native binaries.
-The reason we are using Go and not C/C++ or Java is the recent popularity of it for writing highly concurrent programs.
-Popular open-source projects like \emph{Kubernetes}~\footnote{\url{https://kubernetes.io/}}, \emph{Docker}~\footnote{\url{https://www.docker.com/}} or \emph{Prometheus}~\footnote{\url{https://prometheus.io/}} are all written in Go.
-In the official documentation of Go it claims: ``Go is expressive, concise, clean, and efficient.
-Its concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines, while its novel type system enables flexible and modular program construction.''~\cite{goDocs}
-So, Go tries to solve some of the problems that arise when writing concurrent software application by their design of language.
-% TODO: How does Go want to archive this?
+The increasing demand for parallel programs has also lead to the development of new programming languages which specialize on this task.
+The goal of these languages is to ease the usage of parallel programming patterns so developers can use it more often without writing extensively more code.
+Additionally they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in the Outlook section.
+One example for such a language is Go which is a statically typed programming language developed by Google, which compiles down to single native binaries.
+Go recently gained a lot of popularity for the purpose of writing highly concurrent programs.
+Popular open-source projects like \emph{Kubernetes}\footnote{\url{https://kubernetes.io/}}, \emph{Docker}\footnote{\url{https://www.docker.com/}} or \emph{Prometheus}\footnote{\url{https://prometheus.io/}} are all written in Go.
+In the official documentation of Go states: ``[... Go's] concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines [...]''~\cite{goDocs}
 Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} keyword followed by a named or anonymous function.
@@
 Go encourages programmers to use concurrency by \emph{message passing}, which is assumed to be safer and more convenient to use than \emph{shared memory}.
-However, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
-A study conducted by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
-They came to the conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
-Projects written in Go also tend to have more concurrency than projects written in traditional programming languages such as C.~\cite{tu2019go}
-Given all those conclusions it is even more important for Go developers to catch concurrency bugs effectively.
+For this purpose Go introduces a concept called \emph{channels} where typed messages can be passed through like a pipe.
+Beside the message passing however, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
+Even though Go is designed for programming concurrent programs, this does not mean that Go programs are free of concurrency bugs as we will see in \Cref{sct:outlook}.
+Projects written in Go also tend to have more concurrency than projects written in traditional programming languages such as C~\cite{tu2019go}.
+So it is even more important for Go developers to catch concurrency bugs effectively and as early as possible.
 Together with the increasing popularity of the language, the demand for production ready tools to catch concurrency bugs is very high, which is why this paper evaluates existing techniques in the context of the Go programming language.
 
-\subsection{Table of Contents}
+\subsection{Structure}
 
-\Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes: Deadlocks, Data races and Atomicity and Order violations.
-\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs, reliably reproduce them by utilizing record and replay with the tool \emph{rr} and detect data races dynamically with the tool \emph{ThreadSanitizer}.
-\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically by manipulating the thread scheduler and delta debugging.
-And \Cref{sct:static} finally covers how to detect concurrency bugs with static code analysis by sequentialization and model-checking.
-In the end there is a conclusion with a short comparison of all techniques and a quick outlook on the possible future of multi-threaded debugging.
+The further structure of this paper continues as follows:
+\Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes.
+\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs, reliably reproduce them.
+\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically in a testing environment.
+\Cref{sct:static} covers how to detect concurrency bugs with static code analysis.
+And finally, \Cref{sct:outlook} will provide a quick outlook on how the examined tools can be used together as well as other promising technologies that might help avoiding concurrency bugs in the future.
 
@@
 \subsection{Deadlocks}
-\begin{lstlisting}[float=h, language=Go, label=lst:deadlockWG, caption=Deadlock caused by waiting for the \emph{WaitGroup} at a wrong location -- based on \cite{tu2019go}]
+
+The most common manifestation of blocking bugs are \emph{deadlocks}, where circular dependencies between resources block the flow of a program.
+
+\begin{lstlisting}[float=h, language=Go, label=lst:deadlockCh, caption={Deadlock caused by misuse of an \emph{unbuffered Channel}}]
 func main() {
-    var group sync.WaitGroup
-    group.Add(len(data))
-    for _, d := range data {
-        go func(d string) {
-            fmt.Printf("Processing: %s\n", d)
-            defer group.Done()
-        }(d)
-        group.Wait() // Causing the bug
-    }
+    data := make(chan int)
+    done := make(chan bool)
+    ch <- 1 // Program already blocks here
+    go func() {
+        process(<-data)
+        done <- true
+    }()
+    <-done
 }
@@
 
-The most commonly manifestation of blocking bugs are \emph{deadlocks}, where circular dependencies between resources block the flow of a program.
-\Cref{lst:deadlockWG} shows one example of such a deadlock in a Go program.
-The problem is a \emph{blocking synchronization} where the \lstinline{group.Wait()} inside the for-loop is causing the block.
-This statement has to be moved outside the for loop to resolve the unintentional blocking and fix the concurrency bug.
-Although the error seems obvious in this case, those small mistakes can quickly happen and can get unrecognized into the production environment if not tested well enough.
-
-\begin{lstlisting}[float=h, language=Go, label=lst:deadlockCh, caption={Deadlock caused by misuse of an \emph{unbuffered Channel}}]
-func main() {
-    ch := make(chan int)
-    pr := make(chan int)
-    ch <- 1 // Program already blocks here
-    go func() {
-        i := <-ch
-        pr <- i
-    }()
-    <-pr
-}
-\end{lstlisting}
-
-A second example of a deadlock that might not be obvious is \Cref{lst:deadlockCh} which uses two unbuffered channels to transfer information between threads.
-The problem here is that without an active listener on an unbuffered channel, any send action will be blocked.
+An example of a deadlock in a Go program that might not be obvious is \Cref{lst:deadlockCh} which uses unbuffered channels to transfer information between threads.
+Go channels can either be buffered or unbuffered.
+With unbuffered channels the sender thread has to wait until a second thread is ready to receive the message.
+With the buffered channel the message gets buffered and the sender thread can continue it's execution immediately as long as the buffer is not full.
+In the first two lines of the main method the two unbuffered channels \lstinline{data} and \lstinline{done} are created to send data to the thread and to receive information when the execution of the thread is done.
+Using the arrow notation data can be either send into a channel (line 4) or data can be received from a channel (line 6)
+The bug in this case is that without an active listener on an unbuffered channel, any send action will be blocked.
 To fix this, one could replace the unbuffered channel with a buffered one, so that the execution flow of the program can continue without blocking.
@@
 \subsection{Data Races}
-\begin{lstlisting}[float=h, language=Go, label=lst:race, caption=Data race -- based on \cite{goRaceDetector}]
+
+Data races are part of the non-blocking concurrency bugs.
+They occur when multiple threads simultaneously try to access a shared variable where at least one access is a write operation~\cite{serebry2009threadsanitizer}.
+
+\begin{lstlisting}[float=h, language=Go, label=lst:race, caption=Data race by concurrently accessing a map]
 func main() {
-	c := make(chan bool)
-	m := make(map[string]string)
+	data := make(map[string]string)
+	done := make(chan bool)
 	go func() {
-		m["1"] = "a" // First conflicting access.
-		c <- true
+		data["1"] = "a" // First conflicting access.
+		done <- true
 	}()
-	m["2"] = "b" // Second conflicting access.
-	<-c
+	data["2"] = "b" // Second conflicting access.
+	<-done
 }
@@
 
-Data races are part of the non-blocking concurrency bugs.
-They occur when multiple threads simultaneously try to access a shared variable where at least one access is a write operation.~\cite{serebry2009threadsanitizer}
-
 \Cref{lst:race} shows an example of a data race that is frequently found.~\cite{serebry2009threadsanitizer}
-The data race happens when ``two threads access a non-thread-safe complex object [e.g. a map] without synchronization.''~\cite{serebry2009threadsanitizer}
-Even though the two threads in this example write to different keys of the map \lstinline{m}, this might cause a corruption of data or even crash the program because the default Go map implementation is not concurrency-aware.
-To fix this, the access to the map needs to be synchronized by a lock for example.
+The data race happens when two threads access a non-thread-safe complex object [e.g. a map] without synchronization~\cite{serebry2009threadsanitizer}.
+Even though the two threads in this example write to different keys of the map \lstinline{data}, this might cause a corruption of data or even crash the program because the default Go map implementation is not concurrency-aware.
+To fix this, the access to the map needs to be synchronized, e.g. by a lock.
 
@@
 Whenever the data field gets updated, so does the length field which is semantically correlated with the data.
-When multiple threads update these variables they need to synchronize not only on each of the variables separately but on all semantically correlated variables at the same time.~\cite{lu2007muvi}
+When multiple threads update these variables they need to synchronize not only on each of the variables separately but on all semantically correlated variables at the same time~\cite{lu2007muvi}.
 
@@
 
-Atomicity and order violations are concurrency bugs where the interleaving of threads violate the programmer's intention of atomicity and order.
-This often happens because:
-``Most programmers think sequentially and therefore they make mistakes easily when writing concurrent programs.''~\cite{lu2008mistakes}
-
+Atomicity and order violations are concurrency bugs where the interleaving of threads violates the programmer's intention of atomicity and order.
+% TODO: cite test and use
 \Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''.
@@
         go func() {
-            defer group.Done()
             sum++ // Not atomic
+            group.Done()
         }()
@@
 
+% TODO: cite load-store
 \Cref{lst:atomicity} shows one example of the infamous ``Load-Store'' bug pattern.
-The programmer assumes that ++ is an atomic operation because it is one literal in Go.
-However, after compilation this is expanded to 3 instructions: LOAD, INCREMENT and finally STORE.
-The thread scheduler could switch the context after any of these instructions what leads to undefined behavior, when multiple threads try to increment the same variable.
+The programmer might have assumed that \lstinline{++} is an atomic operation because it is one literal in Go.
+However, after compilation, this is expanded to 3 instructions: LOAD, INCREMENT and finally STORE.
+The thread scheduler could switch the context after any of these instructions.
+This leads to undefined behavior, when multiple threads try to increment the same variable.
 To fix this bug pattern, the \lstinline{++} operation also needs to be replaced by an atomic operation that does not allow other threads to access the \lstinline{sum} variable while incrementing.
-
-Lu, Park, Seo and Zhou conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs.~\cite{lu2008mistakes}
-One key finding was that:
-``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}
-
-A promising solution to atomicity and order violation bugs is \emph{software transactional memory} (STM) as proposed by Peyton Jones.~\cite{peytonjones2007beautiful}
-It is an alternative to traditional lock-based synchronization where atomic regions get declared explicitly.
-Languages like Haskell provide STM by their design of language but Go can also utilize this mechanism by using external libraries.
 
@@
 % ------------------------------------ %
+% -------------- OUTLOOK ------------- %
+% ------------------------------------ %
+\section{Outlook}
+\label{sct:outlook}
+A study conducted by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
+They came to the conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
+
+Lu, Park, Seo and Zhou conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs~\cite{lu2008mistakes}.
+One key finding was that:
+``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}
+
+A promising solution to atomicity and order violation bugs is \emph{software transactional memory} (STM) as proposed by Peyton Jones~\cite{peytonjones2007beautiful}.
+It is an alternative to traditional lock-based synchronization where atomic regions get declared explicitly.
+Languages like Haskell provide STM by their design of language but Go can also utilize this mechanism by using external libraries.
+
+% ------------------------------------ %
 % ------------ CONCLUSION ------------ %

### 2020-05-31 src/paper.tex
Fix some things

@@
     \item Static code analysis with the methods of sequentialization and model-checking.
-    \item Language and software design to avoid concurrency bugs
+    \item Language and software design to avoid concurrency bugs.
 \end{itemize}
@@
 
-\Cref{fig:classes} shows a taxonomy of concurrency bugs based on the work of Tchamgoue, Kim and Jun.~\cite{tchamgoue2012testing}
+\Cref{fig:classes} shows a taxonomy of concurrency bugs based on the work of Tchamgoue, Kim and Jun~\cite{tchamgoue2012testing}.
 They distinguish between two main categories of concurrency bugs: \emph{Blocking} and \emph{non-blocking}.
@@
 
-\Cref{lst:race} shows an example of a data race that is frequently found.~\cite{serebry2009threadsanitizer}
+\Cref{lst:race} shows an example of a data race that is frequently found~\cite{serebry2009threadsanitizer}.
 The data race happens when two threads access a non-thread-safe complex object [e.g. a map] without synchronization~\cite{serebry2009threadsanitizer}.
@@
 
+Assumed, a concurrency bug that rarely manifest gets deployed into production.
+How can it be found quickly?
+And once reported, how can it be reproduced easily for debugging purpose?
 One of the main techniques for finding concurrency bugs is dynamic code analysis.
-% TODO: Find better word for problematic
-In dynamic code analysis the runtime behavior of a program is observed to detect problematic patterns like unsynchronized variable accesses or deadlocks.
+In dynamic code analysis the runtime behavior of a program is observed to detect problematic patterns like un-synchronized variable accesses or deadlocks.
+Assumed, a bug is reported that does not manifest regularly, this technique can help to find the manifestation of the bug more easily during the debugging process.
+Additionally it could be used to observe deployments in production, so concurrency bugs get reported immediately as they occur.
 
-One popular implementation of this is ``Record and Replay'' where the path of execution of an application is recorded and can later be deterministically replayed.
-This way developers can easily find and fix those bug because, ``detecting concurrency bugs is difficult but once detected; correcting them is somehow an easy job.''~\cite{tchamgoue2012testing}
+\subsection{Record and Replay}
+One popular implementation of this technique is called ``Record and Replay'' where the path of execution of an application is recorded and can later be deterministically replayed.
 To create a recording that contains all information to replay a program exactly the same way it was executed, the recorder needs to keep track of the schedule of the program as well as all variables that might not be deterministically reproducible.
-Furthermore, any source of non-determinism such as the the interaction with the network or graphics card or any call to external libraries need to be recorded.~\cite{lidbury2019sparse}
-Like expected, this creates a lot of computational overhead in CPU time as well as storage.
-This approach also forms the foundation for many other techniques and has many different variants and implementations.
+Furthermore, any source of non-determinism such as the the interaction with the network or graphics card or any call to external libraries need to be recorded~\cite{lidbury2019sparse}.
+This creates a lot of computational overhead in CPU time as well as storage.
+This approach also forms the foundation for many other techniques and has many different variants and implementations~\cite{acm2002}~\cite{lidbury2019sparse}.
 
@@
 One requirement to use such a tool in production is the ease of deployability which means that the operational overhead should be very small.
-There have been approaches with custom hardware or patched operating system kernels to hijack the scheduler, but these setups are quite complex and not easy to deploy.
-One example for a tool that fulfills this requirement is \emph{rr}~\footnote{\url{https://rr-project.org/}} developed by \emph{Mozilla}.
-Their approach is to run the program to observe on a fixed CPU core and track the system calls by using \emph{\_ptrace\_}.
-This way they can record programs on an unmodified Linux system and without compiling instrumentation into the programs code which might affect the concurrency behavior.
-The main performance bottleneck beside the execution on only one CPU core, are the context switches induced by ptrace.~\cite{o2017engineering}
+There have been approaches with custom hardware or patched operating system kernels to hijack the scheduler, but these setups are quite complex and not easy to deploy~\cite{o2017engineering}.
+One example for a tool that fulfills this requirement is \emph{rr}~\cite{mozillarr} developed by \emph{Mozilla}.
+Their approach is to run the program to observe on a fixed CPU core and track the system calls by using \emph{ptrace}, which is a system call provided by the Linux kernel to inspect and observe processes.
+This way they can record programs on an unmodified Linux system, which increases the deployability of such a setup.
+With this approach they also do not need to modify the program itself with code instrumentation, where code is added to the program at compile time to observe it's runtime and which might affect the concurrency behavior.
+The main performance bottleneck beside the execution on only one CPU core, are the context switches induced by ptrace~\cite{o2017engineering}.
+
 Lidbury and Donaldson~\cite{lidbury2019sparse} examined further ways to reduce the computational overhead by a sparse recording approach.
-By selectively ignoring ``unimportant'' sources of non-determinism they were able to record input and output heavy applications like video games with an acceptable overhead where as \emph{rr} was not able to record at all due to the massive overhead of recording every source of non-determinism.
+By selectively ignoring ``unimportant'' sources of non-determinism, like the exact layout of memory, they were able to record input- and output-heavy applications like video games with a decreased overhead whereas \emph{rr} was not able to record at all due to the massive overhead of recording every source of non-determinism.
 During replay they execute the thread schedule as recorded and mock every source of recorded non-determinism but let the ignored regions of the program execute non-deterministically.
+Ignoring sources of non-determinism does decrease the accuracy in which concurrency bugs can be replayed.
+But for real-world production usage one has to distinguish between the coverage of realistic bug occurrences and the checking of overall correctness of a program.
+In many cases it is more desirable to observe only one critical region of the code where concurrency bugs are either expected or could lead to serious issues than observing the entire program at once.
+The right balance between coverage and overhead has to be individually evaluated by the safety and speed requirements of a program.
+E.g.
+for a program that is used in a critical system like a banking environment, the developers should aim for a much higher coverage than for a regular web server, where speed is often more important than the overall correctness.
 
-Record and replay could therefore be a great tool to monitor applications in production.
-Once a bug occurs and is either reported by a user or by the program itself, the programmer could just replay the thread schedule to reproduce, find and therefore fix the bug quickly.
+Sparse Record and replay provides the ability to make these decisions and therefore adjust the needs to monitor applications in production.
+Once a bug occurs and is either reported by a user or by the program itself, the programmer could just replay the thread schedule to reproduce, find and therefore fix the bug quickly because, ``detecting concurrency bugs is difficult but once detected; correcting them is somehow an easy job''~\cite{tchamgoue2012testing}.
 Additionally, this technique can help to reproduce all kind of concurrency bugs, no matter if they are blocking or non-blocking.
 For programs that do not heavily depend on parallelism this approach is also very feasible.
-The tool \emph{rr} can record such programs with a performance overhead below the factor of two.
-However, for programs that heavily rely on parallelism the overhead to record and store the thread schedule and all necessary parameters can decrease the performance by more than a factor of 30.~\cite{o2017engineering}
-Especially for multi-threaded server applications that expect a high workload and where speed is crucial for the business, this approach is not feasible in production, yet.
+The tool \emph{rr} can record such programs with a performance overhead below the factor of two~\cite{o2017engineering}.
+However, for programs that heavily rely on parallelism the overhead to record and store the thread schedule and all necessary parameters can decrease the performance by more than a factor of 30~\cite{o2017engineering}.
+Especially for multi-threaded server applications that expect a high workload and where speed is crucial, this approach is not feasible in production, yet.
 And even if the workload is low but the application is time-sensitive for example because of protocol timeouts, it is not applicable without extending the acceptable timeouts.
 
+\subsection{Data Race Detection}
 Another set of tools for dynamic code analysis are dynamic data race detectors.
 As the name implies, these tools detect data races but because atomicity and order violations are often correlated, these tools can detect a variety of non-blocking concurrency bugs.
-There are multiple approaches for data race detection but in this case we refer to on-the-fly and post-mortem data race detection which are often referred to as dynamic.~\cite{serebry2009threadsanitizer}
-One data race detector that is mainly used for Go programs is the \emph{ThreadSanitizer} developed by Google.
+There are multiple approaches for data race detection but in this case we refer to on-the-fly and post-mortem data race detection which are typically referred to as dynamic~\cite{serebry2009threadsanitizer}.
+One data race detector that is mainly used for Go programs is the \emph{ThreadSanitizer} in version 2~\cite{threadSanitizer} developed by Google.
 It is built in the Golang compiler tool-chain and can be toggled with a simple compiler flag which makes it very easy to use.
-``The tool performs compile-time instrumentation of the source program, in which all (atomic and non-atomic) accesses to potentially shared locations, as well as fence operations, are instrumented with calls into a statically linked run-time library.''~\cite{lidbury2019sparse}
+The tool performs compile-time instrumentation of the source program, in which all (atomic and non-atomic) accesses to potentially shared locations, as well as fence operations, are instrumented with calls into a statically linked run-time library~\cite{lidbury2019sparse}.
 To distinguish between synchronized accesses and data races during runtime, the compiler also has to inject dynamic annotations into the program.
 With these annotations the data race detector understands the synchronization primitives used by the programming language.
-When the program along with the race detector gets executed, the data race detector ``observes the program execution as a sequence of events.''~\cite{serebry2009threadsanitizer}
-To detect a data race, there are two algorithms that are mainly used among all detectors: \emph{happens-before} and \emph{lockset}.
-Locking events are Write and Read Locks / Unlocks.
-Happens-before events are Signal and Wait.
-The \emph{ThreadSanitizer} uses an hybrid approach by combining both algorithms into one and observing events of both categories.
+When the program along with the race detector gets executed, the data race detector observes the program execution as a sequence of events~\cite{serebry2009threadsanitizer}.
+Each memory access is recorded and when receiving events for memory access it can report a potential race based on a \emph{state machine algorithm}~\cite{serebry2011llvm}.
 
-Performance-wise the race detection is unfortunately not better than the record and replay approach.
-The official race detector article on the Go website claims: ``The cost of race detection varies by program, but for a typical program, memory usage may increase by 5-10x and execution time by 2-20x.''~\cite{goRaceDetector}
-Which is in fact better than the original 20x-50x evaluated by Serebryany and Iskhodzhanov in 2009.~\cite{serebry2009threadsanitizer}
-However, a possible improvement could be to ignore certain uncritical regions and just observe hotspots where data races are suspected.
+Performance-wise, race detection is unfortunately not better than the record and replay approach.
+According to the official Go website: ``The cost of race detection varies by program, but for a typical program, memory usage may increase by 5-10x and execution time by 2-20x''~\cite{goRaceDetector}.
+Which is in fact a massive improvement from \emph{ThreadSanitizer} version 1 with an overhead of 20x-50x, which was evaluated by Serebryany and Iskhodzhanov in 2009~\cite{serebry2009threadsanitizer}.
+However, to improve this even further one possibility could be to ignore certain uncritical regions and just observe hotspots where data races are suspected, like in the sparse record and replay approach.
+The overall coverage of data races would decrease but the developers could have the choice balance speed and correctness according to the requirements.
 
-Beside their ease of use, one of the main disadvantages of dynamic analysis tools is their coverage, that `` [...] is typically small since only a few executions are explored.''~\cite{qadeer2004kiss}
-Additionally, they depend on the manifestation and especially the recognition of these bugs to report them correctly.
+Beside their ease of use, one of the main disadvantages of dynamic analysis tools is their bug coverage, that is typically small since only a few executions are explored~\cite{qadeer2004kiss}.
+These tools just observe and do not actively try to enforce a manifestation of bugs, so not every thread interleaving might be explored.
+Additionally, they depend on the the recognition of these bugs to report them correctly.
+While this is good to deploy them into a production setup to monitor an application, it is also desirable to prevent concurrency bugs before they can even get into production.
 
@@
 
-Extensive testing has proven to minimize the bugs of software in production.\cite{makinen2014testing}
-However, traditional testing mostly covers sequential errors and cannot detect concurrency bugs effectively.\cite{lu2008mistakes}
-This lies in the nature of these bugs themself.
+To prevent concurrency bugs to get into production and eventually causing damage, the software should be tested for those bugs before.
+Extensive testing has proven to minimize the bugs of software in production~\cite{makinen2014testing}.
+However, traditional testing mostly covers sequential errors and cannot detect concurrency bugs effectively~\cite{lu2008mistakes}.
+This lies in the nature of these bugs themselves.
 When only certain thread interleavings trigger the bug, the tests could complete successfully almost every time without ever finding a manifestation.
-So, new concurrency-aware tests are needed to find concurrency bugs reliably.
+For lack of established terms we will use the term \emph{concurrency-aware testing} to describe the process of automatically finding concurrency bugs in a testing environment by intentionally manipulating the path of execution.
 
-One early attempt to test programs for concurrency bugs was to inject delays at random points in the program code, but it is yet unclear if this really gives an acceptable bug coverage.~\cite{lu2008mistakes}
+One early attempt to test programs for concurrency bugs was to inject delays at random points in the program code~\cite{bron2005coverage}, but it is yet unclear if this heuristic really gives an acceptable bug coverage~\cite{lu2008mistakes}.
 Furthermore, the tests should try to reduce the perturbation on the execution, so the software can be evaluated the same way as it would run in a regular deployment in production.
 
-Another method of concurrency-aware testing has been proposed by Choi and Zeller.\cite{acm2002}
-They implemented a tool called ``DEJAVU'' for the \emph{Jalapeño} Java Virtual Machine but the concept is also applicable to tools that work with Go or any other programming language.
-They make use of a technique called \emph{Delta debugging}, which describes the process of narrowing down the spot where a failure is introduced by going back and forth between working and non-working conditions, in this case working and non-working thread schedules.~\cite{zeller2002delta}
+Another method of concurrency-aware testing has been proposed by Choi and Zeller~\cite{acm2002}.
+They implemented a tool called ``DEJAVU''~\cite{acm2002} for the \emph{Jalapeño} Java Virtual Machine but the concept is also applicable to tools that work with Go or any other programming language.
+They make use of a technique called \emph{Delta debugging}, which describes the process of narrowing down the spot where a failure is introduced by going back and forth between working and non-working conditions, in this case working and non-working thread schedules~\cite{zeller2002delta}.
 \Cref{fig:testing} shows a modified model of the \emph{DEJAVU} approach, suited for Go programs by using rr instead of of \emph{DEJAVU} for the recording and replay of the thread schedule.
@@
 
-While this technique cannot be used in a productive deployment setup, it is intended to extend existing tests that for example run inside a CI pipeline.
 One major drawback however, is the number of thread interleavings that can grow exponentially with the number of executed threads.
@@
 Lu et al.
-have found in their study of real-world concurrency bugs that ``Almost 96\% of the examined concurrency bugs are guaranteed to manifest if certain partial order between 2 threads is enforced.''~\cite{lu2008mistakes}
-Which means that ``Pairwise testing on concurrent program threads can expose most concurrency bugs, and greatly reduce the testing complexity.''~\cite{lu2008mistakes}
+have found in their study of real-world concurrency bugs that almost all concurrency bugs were guaranteed to manifest if a certain partial order between two threads is enforced~\cite{lu2008mistakes}.
+Which means that pairwise testing on concurrent program threads can expose most concurrency bugs, and greatly reduce the testing complexity~\cite{lu2008mistakes}.
 To increase the coverage of concurrency bugs, this approach could also be combined with a data race detector to dynamically detect data races within certain replays of a thread schedule.
 This way the test cases can be minimized to semantical correctness checks, which decreases the operational overhead for such a setup.
-The setup of such a system also comes with a relatively low operational overhead because once such a system is build, it could be easily embedded within the CI pipelines of a project.
+
+Concurrency-aware testing is very promising in the regard of extending an existing testing pipeline.
+This way concurrency bugs could be detected even before the software is deployed into production.
+The overhead can be minimized by optimizing the thread schedule generator and by ignoring most uncritical regions of the code.
+However, the coverage depends on the quality of the tests that also need to be aware of multi-variable semantic data races.
 
@@
 
-Static code analysis describes the technique of predicting the behavior of a program by analyzing the source code without compiling nor executing it.
+Even though concurrency-aware testing can prevent some concurrency bugs to get in production, it would be even better if those bugs could be already detected during or before compile time.
+For this purpose static code analysis can be used.
+Static code analysis describes the technique of predicting the behavior of a program before it is run.
 
@@
 But model checking for concurrent programs is complex due to the number of possible thread interleavings that can grow exponentially with the number of executed threads.
-Therefore, static data race detectors for example are unfeasible for very large code bases.~\cite{serebry2009threadsanitizer}
+Therefore, static data race detectors for example are unfeasible for very large code bases~\cite{serebry2009threadsanitizer}.
 
-One implementation of static code analysis to detect concurrency bugs is sequentialization as proposed by Qadeer and Wu.~\cite{qadeer2004kiss}
+One implementation of static code analysis to detect concurrency bugs is sequentialization as proposed by Qadeer and Wu~\cite{qadeer2004kiss}.
 Their approach is to transform a parallel program into a sequential one and simulate a large subset of the behavior of the parallel program on the sequential one.
@@
 Although unfeasible for detecting data races in large code bases, there are multiple implementations of static deadlock-detectors for concurrent programs.
-One concrete static deadlock detector for the Go programming language is called \emph{Godel Checker}.~\cite{godelChecker}
+One concrete static deadlock detector for the Go programming language is called \emph{Godel Checker}~\cite{godelChecker}.
 It uses that Go's message passing concurrency mechanisms by channels is ``[...] greatly inspired by advances in formal languages for concurrency theory known as process calculi.''~\cite{lange2018verification}
@@
 The computational overhead for this approach is also surprisingly low.
-A program with 18 kLOC can be checked within minutes.~\cite{lange2018verification}
+A program with 18 kLOC can be checked within minutes~\cite{lange2018verification}.
 Unfortunately this technique has quite a few limitations.
@@
 
-Concurrency-aware testing is very promising in the regard of extending an existing testing pipeline.
-This way concurrency bugs could be detected even before the software is deployed into production.
-The overhead can be minimized by optimizing the thread schedule generator and by ignoring most uncritical regions of the code.
-However, the coverage depends on the quality of the tests that also need to be aware of multi-variable semantic data races.
-
 Static code analysis is only feasible in small code bases, due to the exponentially growing thread interleavings that need to be explored, which requires a lot of computational overhead.

### 2020-05-31 src/paper.tex
Fix some more things

@@
 In this paper we evaluate different techniques and tools that are currently available to effectively find, fix and avoid concurrency bugs.
-The requirements for these tools are that they need to be easy to use, they should bring only a minimal computational and storage overhead and they should have a high coverage with minimal false-positive reports.
+These tools should be easy to use, they should bring only a minimal computational and storage overhead and they should have a high coverage with minimal false-positive reports.
 In short: These tools should be production-ready and applicable to real-world software projects.
@@
     \item Static code analysis with the methods of sequentialization and model-checking.
-    \item Language and software design to avoid concurrency bugs.
+    \item Language and software design to avoid concurrency bugs when writing software.
 \end{itemize}
@@
 Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
-To utilize the full performance of the given hardware, programmers need to write their software with a high degree of parallelism.
+To utilize the full performance of the given hardware, developers need to write their software with a high degree of parallelism.
 In contrast to sequential programs which only have one deterministic path through the program code, concurrent programs have multiple paths of execution that are executed independently from each other.
@@
 \begin{quote}
-``But Heisenbugs may elude a bugcatcher for years of execution. Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.''\cite{gray1986computers}
+``[...] Heisenbugs may elude a bugcatcher for years of execution. Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.''\cite{gray1986computers}
 \end{quote}
 
-Heisenbugs are bugs that only occur with very specific thread interleavings that might not occur when a bugcatcher evaluates the program.
+Heisenbugs are bugs that only manifest within very specific thread interleavings that might not occur when a bugcatcher evaluates the program.
 The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix and avoid them~\cite{tu2019go}.
@@
 The goal of these languages is to ease the usage of parallel programming patterns so developers can use it more often without writing extensively more code.
-Additionally they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in the Outlook section.
+Additionally they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in \cref{sct:outlook}.
 One example for such a language is Go which is a statically typed programming language developed by Google, which compiles down to single native binaries.
@@
 Popular open-source projects like \emph{Kubernetes}\footnote{\url{https://kubernetes.io/}}, \emph{Docker}\footnote{\url{https://www.docker.com/}} or \emph{Prometheus}\footnote{\url{https://prometheus.io/}} are all written in Go.
-In the official documentation of Go states: ``[... Go's] concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines [...]''~\cite{goDocs}
+The official documentation of Go states: ``[... Go's] concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines [...]''~\cite{goDocs}.
 Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} keyword followed by a named or anonymous function.
@@
 
-An example of a deadlock in a Go program that might not be obvious is \Cref{lst:deadlockCh} which uses unbuffered channels to transfer information between threads.
+An example of a deadlock in a Go program that might not be obvious is shown in \cref{lst:deadlockCh}, where unbuffered channels are used to transfer information between threads.
 Go channels can either be buffered or unbuffered.
 With unbuffered channels the sender thread has to wait until a second thread is ready to receive the message.
-With the buffered channel the message gets buffered and the sender thread can continue it's execution immediately as long as the buffer is not full.
-In the first two lines of the main method the two unbuffered channels \lstinline{data} and \lstinline{done} are created to send data to the thread and to receive information when the execution of the thread is done.
-Using the arrow notation data can be either send into a channel (line 4) or data can be received from a channel (line 6)
+With buffered channels the message gets buffered and the sender thread can continue it's execution immediately as long as the buffer is not full.
+In the first two lines of the main method of \cref{lst:deadlockCh}, the two unbuffered channels \lstinline{data} and \lstinline{done} are created to send data to the thread and to receive information when the execution of the thread is done.
+Using the arrow notation, data can either be send into a channel (line 4) or data can be received from a channel (line 6).
 The bug in this case is that without an active listener on an unbuffered channel, any send action will be blocked.
@@
 Another common problem in event-driven concurrent programs are \emph{blocking operations} like filesystem operations that are executed inside an event-handler.
-These ``can penalize and even paralyze the entire program execution.''~\cite{tchamgoue2012testing}
+These can penalize and even paralyze the entire program execution~\cite{tchamgoue2012testing}.
 
@@
 \Cref{lst:race} shows an example of a data race that is frequently found~\cite{serebry2009threadsanitizer}.
-The data race happens when two threads access a non-thread-safe complex object [e.g. a map] without synchronization~\cite{serebry2009threadsanitizer}.
+The data race happens when two threads access a non-thread-safe complex object (e.g. a map) without synchronization.
 Even though the two threads in this example write to different keys of the map \lstinline{data}, this might cause a corruption of data or even crash the program because the default Go map implementation is not concurrency-aware.
@@
 
-Assumed, a concurrency bug that rarely manifest gets deployed into production.
+Assumed, a concurrency bug that rarely manifests, gets deployed into production.
 How can it be found quickly?
@@
 There have been approaches with custom hardware or patched operating system kernels to hijack the scheduler, but these setups are quite complex and not easy to deploy~\cite{o2017engineering}.
-One example for a tool that fulfills this requirement is \emph{rr}~\cite{mozillarr} developed by \emph{Mozilla}.
+One example for a tool that fulfills the requirements is \emph{rr}~\cite{mozillarr} developed by \emph{Mozilla}.
 Their approach is to run the program to observe on a fixed CPU core and track the system calls by using \emph{ptrace}, which is a system call provided by the Linux kernel to inspect and observe processes.
@@
 Sparse Record and replay provides the ability to make these decisions and therefore adjust the needs to monitor applications in production.
-Once a bug occurs and is either reported by a user or by the program itself, the programmer could just replay the thread schedule to reproduce, find and therefore fix the bug quickly because, ``detecting concurrency bugs is difficult but once detected; correcting them is somehow an easy job''~\cite{tchamgoue2012testing}.
+Once a bug occurs and is either reported by a user or by the program itself, the developer could just replay the thread schedule to reproduce, find and therefore fix the bug quickly because, ``detecting concurrency bugs is difficult but once detected; correcting them is somehow an easy job''~\cite{tchamgoue2012testing}.
 Additionally, this technique can help to reproduce all kind of concurrency bugs, no matter if they are blocking or non-blocking.
@@
 One data race detector that is mainly used for Go programs is the \emph{ThreadSanitizer} in version 2~\cite{threadSanitizer} developed by Google.
-It is built in the Golang compiler tool-chain and can be toggled with a simple compiler flag which makes it very easy to use.
+It is build into the Golang compiler tool-chain and can be toggled with a simple compiler flag which makes it very easy to use.
 The tool performs compile-time instrumentation of the source program, in which all (atomic and non-atomic) accesses to potentially shared locations, as well as fence operations, are instrumented with calls into a statically linked run-time library~\cite{lidbury2019sparse}.
@@
 However, to improve this even further one possibility could be to ignore certain uncritical regions and just observe hotspots where data races are suspected, like in the sparse record and replay approach.
-The overall coverage of data races would decrease but the developers could have the choice balance speed and correctness according to the requirements.
+The overall coverage of data races would decrease but the developers could have the choice of balancing speed and correctness according to the requirements.
 
-Beside their ease of use, one of the main disadvantages of dynamic analysis tools is their bug coverage, that is typically small since only a few executions are explored~\cite{qadeer2004kiss}.
+The main advantage of the dynamic approach is that it theoretically could be deployed to a real production environment.
+For programs that are mainly event-driven this would give the ability to observe the real-world behavior in a reproducible way.
+However, one of the main disadvantages of dynamic analysis tools is their bug coverage, that is typically small since only a few executions are explored~\cite{qadeer2004kiss}.
 These tools just observe and do not actively try to enforce a manifestation of bugs, so not every thread interleaving might be explored.
 Additionally, they depend on the the recognition of these bugs to report them correctly.
-While this is good to deploy them into a production setup to monitor an application, it is also desirable to prevent concurrency bugs before they can even get into production.
+Programs that heavily rely on multi-threaded execution like web-servers can also become very inefficient or could even fail due to timeout limitations.
+To avoid the overhead in the production environment and to find concurrency bugs in programs that could not be observed easily, it is desirable to prevent concurrency bugs before they can even get into production.
 
@@
 One concrete static deadlock detector for the Go programming language is called \emph{Godel Checker}~\cite{godelChecker}.
-It uses that Go's message passing concurrency mechanisms by channels is ``[...] greatly inspired by advances in formal languages for concurrency theory known as process calculi.''~\cite{lange2018verification}
+It uses that Go's message passing concurrency mechanisms by channels is inspired by advances in formal languages for concurrency theory known as process calculi~\cite{lange2018verification}.
 The \emph{Godel Checker} tool extracts the main concurrency characteristics of the program into a mathematical model called \emph{Behavioural Types}.
@@
 
+Static code analysis is only feasible in small code bases, due to the exponentially growing thread interleavings that need to be explored, which requires a lot of computational overhead.
+Additionally multi-variable accesses of only semantically correlated variables need manual annotations or configuration.
+``Unfortunately, existing techniques cannot effectively extract such semantic correlations.
+Traditional compiler analysis cannot catch them, because many correlated variables are just semantically correlated and do not necessarily have data dependencies [...]''~\cite{lu2007muvi}.
+
 % ------------------------------------ %
@@
 \label{sct:outlook}
-A study conducted by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
-They came to the conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
 
-Lu, Park, Seo and Zhou conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs~\cite{lu2008mistakes}.
-One key finding was that:
-``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}
+All the techniques shown can find concurrency bugs with different coverage based on the overhead one is willing to pay.
+As seen, there is not a single tool that fulfills all the requirements for an optimal production usage.
+Either the overhead is very big or the coverage is quite small.
+To get the most out of these tools, a combined approach seems to make sense to get the most coverage with the least overhead.
+
+For the goal of having no concurrency bugs at all in a production deployment with nearly no overhead, it would be desirable to have techniques that make it impossible to even write incorrect concurrent programs.
+New languages and language features have been developed for this task.
+Go as presented before is one of those languages that tries to make the use of concurrent patterns safer.
+One of their main approaches is to encourage developers to synchronize threads via easy to use \emph{message passing}, namely channels.
+A study conducted by Tu, Liu, Song and Zhan~\cite{tu2019go} analyzed popular real-world Go projects and their concurrency bugs.
+They came to the surprising conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
+So, synchronization via \emph{message passing} does not seem to be the right solution to decrease concurrency bugs.
 
@@
 It is an alternative to traditional lock-based synchronization where atomic regions get declared explicitly.
+This way STM is one possibility to ensure that there are no side-effects of atomic operations and no data races.
 Languages like Haskell provide STM by their design of language but Go can also utilize this mechanism by using external libraries.
+Lu, Park, Seo and Zhou~\cite{lu2008mistakes} conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs.
+One of their key finding was that most of the examined non-deadlock concurrency bugs were covered by atomicity-violation or order-violation patterns.
+Their conclusion is that using STM could have avoided about one third of the examined concurrency bugs.
+
+Combining all this with the ongoing trend of using micro-service architectures over big monoliths, these tools could get a lot more useful because the code bases of each service tend to get smaller.
+This means that the overhead of instrumenting, testing or model-checking one service gets smaller in absolute numbers.
+This way teams can decide more easily which applications need to have a higher concurrency bug coverage and which applications can be ignored.
+Record and replay could also be deployed to just some replicas of an instance, reducing the impact on speed whilst collecting useful information to optimize the whole code base.
 
@@
 
-While having multi-threaded applications is convenient and also necessary in today's world of computation, the non-determinism makes it hard to find, reproduce and therefore fix concurrency bugs.
 Although there is a variety of tools that try to ease the process of multi-threaded debugging, as shown, none of them really fulfills all requirements for a production-ready drop-in solution.
-
-Dynamic code analysis has the main disadvantage that it can cost a lot of overhead when monitoring is active in a production environment.
-Especially programs that heavily rely on multi-threaded execution like web-servers  can become very inefficient or could even fail due to timeout limitations.
-Dynamic data race detectors could eventually be optimized to an acceptable overhead by ignoring most of the assumed-to-be-safe code regions, but this also drastically lowers the overall coverage.
-Beside that, ``data-race free does not mean concurrency bug free.''~\cite{lu2008mistakes}
-The main advantage of the dynamic approach is that it theoretically could be deployed to a real production environment.
-For programs that are mainly event-driven this would give the ability to observe the real-world behavior in a reproducible way.
-
-Static code analysis is only feasible in small code bases, due to the exponentially growing thread interleavings that need to be explored, which requires a lot of computational overhead.
-Additionally multi-variable accesses of only semantically correlated variables need manual annotations or configuration.
-``Unfortunately, existing techniques cannot effectively extract such semantic correlations.
-Traditional compiler analysis cannot catch them, because many correlated variables are just semantically correlated and do not necessarily have data dependencies, [...]''~\cite{lu2007muvi}
-Beside that there are some promising tools for statically finding deadlocks by translating programs into a mathematical model.
-
-Concluding this paper, the evaluation of the different approaches shows that there is not one tool to solve all problems that arise when writing concurrent programs.
 But there are multiple techniques that can be combined to at least get some support in debugging multi-threaded applications.
-
-However, for modern micro-service architectures that get more and more popular, these tools could actually be useful because the code bases tend to get smaller.
-That means that the overhead of instrumenting, testing or model-checking these applications gets smaller in absolute numbers.
-Record and replay could also be deployed to just some replicas of an instance, reducing the impact on crucial business speed whilst collecting useful information to optimize the whole code base.
-
-But then again, detecting and reporting concurrency bugs is just the first step in fixing them.
-Developers need to understand the reporting and need to distinguish between real bugs and false-positives.
-So, ``There is no silver bullet for fixing concurrency bugs.''~\cite{lu2008mistakes}
+In all of this the developers always have to decide if safety or speed is more important for their application and adjust the tools to these needs.
 

### 2020-05-31 src/paper.tex
Fix small mistakes

@@
 In this paper we evaluate different techniques and tools that are currently available to effectively find, fix and avoid concurrency bugs.
-These tools should be easy to use, they should bring only a minimal computational and storage overhead and they should have a high coverage with minimal false-positive reports.
+These tools should be easy to use, they should operate with a minimal computational and storage overhead and they should have a high coverage with minimal false-positive reports.
 In short: These tools should be production-ready and applicable to real-world software projects.
@@
     \item Dynamic code analysis with record and replay and dynamic data race detection.
-    \item Concurrency-aware testing with a combined approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
+    \item Concurrency-aware testing with the approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
     \item Static code analysis with the methods of sequentialization and model-checking.
-    \item Language and software design to avoid concurrency bugs when writing software.
+    \item Programming language and software design to avoid concurrency bugs when writing software.
 \end{itemize}
@@
 
-This paper will evaluate some of these techniques and tools and how they can be used to ease the process of debugging concurrent bugs.
+In this paper we evaluate some of these techniques and tools and how they can be used to ease the process of debugging concurrent bugs.
 The main focus is to detect bugs efficiently in production, which means that developers should be able to fix bugs with a minimum amount of work and time.
@@
 \subsection{The Go Programming Language}
-The increasing demand for parallel programs has also lead to the development of new programming languages which specialize on this task.
-The goal of these languages is to ease the usage of parallel programming patterns so developers can use it more often without writing extensively more code.
-Additionally they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in \cref{sct:outlook}.
-One example for such a language is Go which is a statically typed programming language developed by Google, which compiles down to single native binaries.
+The increasing demand for parallel programs has also lead to the development of new programming languages which specialize in this task.
+The goal of these languages is to ease the usage of parallel programming patterns, so developers can use them more often without writing extensively more code.
+Additionally, they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in \cref{sct:outlook}.
+One example for such a language is Go which is a statically typed programming language developed by Google, that compiles down to single native binaries.
 Go recently gained a lot of popularity for the purpose of writing highly concurrent programs.
-Popular open-source projects like \emph{Kubernetes}\footnote{\url{https://kubernetes.io/}}, \emph{Docker}\footnote{\url{https://www.docker.com/}} or \emph{Prometheus}\footnote{\url{https://prometheus.io/}} are all written in Go.
+Popular open-source projects like \emph{Kubernetes}\footnote{\url{https://kubernetes.io/}}, \emph{Docker}\footnote{\url{https://www.docker.com/}} and \emph{Prometheus}\footnote{\url{https://prometheus.io/}} are all written in Go.
 The official documentation of Go states: ``[... Go's] concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines [...]''~\cite{goDocs}.
@@
 \Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes.
-\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs, reliably reproduce them.
+\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs and reliably reproduce them.
 \Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically in a testing environment.
@@
 Atomicity and order violations are concurrency bugs where the interleaving of threads violates the programmer's intention of atomicity and order.
-% TODO: cite test and use
-\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''.
+\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''~\cite{farchi2003patterns}.
 The programmer's intention is to check if a variable is not \lstinline{nil} and then use this variable.
@@
 
-% TODO: cite load-store
-\Cref{lst:atomicity} shows one example of the infamous ``Load-Store'' bug pattern.
+\Cref{lst:atomicity} shows one example of the infamous ``Load-Store'' bug pattern~\cite{farchi2003patterns}.
 The programmer might have assumed that \lstinline{++} is an atomic operation because it is one literal in Go.
@@
 Programs that heavily rely on multi-threaded execution like web-servers can also become very inefficient or could even fail due to timeout limitations.
-To avoid the overhead in the production environment and to find concurrency bugs in programs that could not be observed easily, it is desirable to prevent concurrency bugs before they can even get into production.
+To avoid the overhead in the production environment and to find concurrency bugs in programs that could not be observed easily, it is desirable to find concurrency bugs before they can even get into production.
 
@@
 Concurrency-aware testing is very promising in the regard of extending an existing testing pipeline.
-This way concurrency bugs could be detected even before the software is deployed into production.
+This way concurrency bugs could be detected even before the software gets deployed into production.
 The overhead can be minimized by optimizing the thread schedule generator and by ignoring most uncritical regions of the code.
@@
 But there are multiple techniques that can be combined to at least get some support in debugging multi-threaded applications.
-In all of this the developers always have to decide if safety or speed is more important for their application and adjust the tools to these needs.
+In all of this the developers always have to decide if safety and overall correctness or speed is more important for their application and adjust the tools to these needs.
 

### 2020-06-28 src/paper.tex
Fix unclear parts

@@
 \begin{abstract}
-Concurrency bugs are often hard to find and therefore hard to fix.
+Bugs that are caused by the parallel execution of a program, so called
+concurrency bugs are often hard to find and therefore hard to fix.
 In this paper we evaluate different techniques and tools that are currently available to effectively find, fix and avoid concurrency bugs.
-These tools should be easy to use, they should operate with a minimal computational and storage overhead and they should have a high coverage with minimal false-positive reports.
-In short: These tools should be production-ready and applicable to real-world software projects.
+These tools are evaluated by their ease of use, their computational and storage overhead, their concurrency bug coverage and their potential amount of false-positive reports.
+The goal is to find tools that are production-ready and applicable to real-world software projects.
 The techniques evaluated are:
 \begin{itemize}
-    \item Dynamic code analysis with record and replay and dynamic data race detection.
-    \item Concurrency-aware testing with the approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
+    \item Dynamic code analysis in form of record and replay and dynamic data race detection.
+    \item Concurrency-aware testing by evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
     \item Static code analysis with the methods of sequentialization and model-checking.
-    \item Programming language and software design to avoid concurrency bugs when writing software.
+    \item Programming language design patterns to avoid concurrency bugs when writing software.
 \end{itemize}
@@
 
-\subsection{Heisenbugs}
-
 \begin{quote}
@@
 
+\subsection{Heisenbugs}
+
 % Motivation
 Today's computer architectures rely more and more on parallel computing.
-Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
+Single-core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
 To utilize the full performance of the given hardware, developers need to write their software with a high degree of parallelism.
-In contrast to sequential programs which only have one deterministic path through the program code, concurrent programs have multiple paths of execution that are executed independently from each other.
+In contrast to sequential programs which only follow one predictable path through the program code, concurrent programs can have multiple paths that are executed independently from each other.
+Using threads that branch off from the main program path, it is possible to execute operations in parallel.
 The order in which these threads are executed is determined by the operating system's scheduler.
@@
 Such bugs can be very challenging to debug because they are not easily reproducible~\cite{tu2019go}.
-Due to this complexity, it is no surprise that bugs that are related to concurrency and parallel program execution occur quite often.
-But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread interleavings appear, cannot be reliably reproduced without knowing the schedule it appeared in.
+Due to this complexity, it is no surprise that bugs that are related to concurrency and parallel program execution occur quite often~\cite{lu2008mistakes}.
+But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread interleavings appear cannot be reliably reproduced without knowing the schedule it appeared in.
 
@@
 Heisenbugs are bugs that only manifest within very specific thread interleavings that might not occur when a bugcatcher evaluates the program.
-The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix and avoid them~\cite{tu2019go}.
-Thus, numerous tools and techniques have been developed that propose different solutions to solve the problem of concurrency bugs and accelerate the debugging process.
+The complexity and uncertainty of concurrency bugs is also the reason for studies~\cite{tu2019go} that examine how these bugs occur and explore different ways on how to detect, fix and avoid them.
+Thus, numerous tools and techniques have been developed that propose different solutions to solve the problem of concurrency bugs or accelerate the debugging process.
 
@@
 The goal of these languages is to ease the usage of parallel programming patterns, so developers can use them more often without writing extensively more code.
-Additionally, they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in \cref{sct:outlook}.
-One example for such a language is Go which is a statically typed programming language developed by Google, that compiles down to single native binaries.
+Additionally, they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in \Cref{sct:outlook}.
+One example for such a language is Go which is a statically typed programming language developed by Google, that compiles down to native binaries~\cite{goDocs}.
 Go recently gained a lot of popularity for the purpose of writing highly concurrent programs.
@@
 Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} keyword followed by a named or anonymous function.
-The Go runtime maps these goroutines to normal Kernel threads during execution.
+The Go runtime maps these goroutines to operating system threads during execution.
 Go encourages programmers to use concurrency by \emph{message passing}, which is assumed to be safer and more convenient to use than \emph{shared memory}.
-For this purpose Go introduces a concept called \emph{channels} where typed messages can be passed through like a pipe.
+For this purpose Go is using a concept called \emph{channels} where typed messages can be passed through like a UNIX pipeline.
 Beside the message passing however, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
@@
 So it is even more important for Go developers to catch concurrency bugs effectively and as early as possible.
-Together with the increasing popularity of the language, the demand for production ready tools to catch concurrency bugs is very high, which is why this paper evaluates existing techniques in the context of the Go programming language.
+Together with the increasing popularity of the language, the demand for production-ready tools to catch concurrency bugs is very high, which is why this paper evaluates existing techniques in the context of the Go programming language.
 
@@
 \Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs and reliably reproduce them.
-\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically in a testing environment.
+\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically in a testing environment, where in contrast to a production environment concurrency bugs are explicitly provoked.
 \Cref{sct:static} covers how to detect concurrency bugs with static code analysis.
@@
 Blocking bugs are bugs where the execution of a program is unintentionally blocked and the program cannot terminate.
-% TODO: Meh, don't know if that is right..
-The manifestation of these kind of bugs is very noticeable because the whole application freezes and has to be shutdown before the computation can continue.
-Non-blocking bugs are often harder to find because they can occur even when the termination of a program was successful but the result is wrong.
+The manifestation of these kind of bugs is very noticeable because the part of a program in which the deadlock appears in freezes and has to be restarted before the computation can continue.
+Non-blocking bugs are often harder to find because they can occur even when the execution of a program does not block but the calculations are wrong.
 This can also lead to a cascade of bugs where the root cause is a non-blocking concurrency bug which might not be obvious.
@@
 In the first two lines of the main method of \cref{lst:deadlockCh}, the two unbuffered channels \lstinline{data} and \lstinline{done} are created to send data to the thread and to receive information when the execution of the thread is done.
-Using the arrow notation, data can either be send into a channel (line 4) or data can be received from a channel (line 6).
+Using the arrow notation, data can either be sent into a channel (line 4) or data can be received from a channel (line 6).
 The bug in this case is that without an active listener on an unbuffered channel, any send action will be blocked.
@@
 Another common problem in event-driven concurrent programs are \emph{blocking operations} like filesystem operations that are executed inside an event-handler.
-These can penalize and even paralyze the entire program execution~\cite{tchamgoue2012testing}.
+Because event-handlers have a higher priority than non-event code, these operations can penalize and even paralyze the entire program execution~\cite{tchamgoue2012testing}.
 
@@
 \subsection{Atomicity and Order Violations}
+
+Atomicity and order violations are concurrency bugs where the interleaving of threads violates the programmer's intention of atomicity and order.
+\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''~\cite{farchi2003patterns}.
+The programmer's intention is to check if a variable is not \lstinline{nil} and then use this variable.
+However, due to the thread that was launched before, it could happen that after the \lstinline{if} check in line 7, the thread of the goroutine gets scheduled and the data variable is set to \lstinline{nil}.
+To fix this bug, the check and the usage of the variable need to become an atomic operation to enforce the order of execution.
+
 \begin{lstlisting}[float=h, language=Go, label=lst:order, caption=Test-and-Use bug pattern -- Order violation]
@@
 \end{lstlisting}
-
-Atomicity and order violations are concurrency bugs where the interleaving of threads violates the programmer's intention of atomicity and order.
-\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''~\cite{farchi2003patterns}.
-The programmer's intention is to check if a variable is not \lstinline{nil} and then use this variable.
-However, due to the thread that was launched before, it could happen that after the \lstinline{if} check in line 7, the thread of the goroutine gets scheduled and the data variable is set to \lstinline{nil}.
-To fix this bug, the check and the usage of the variable need to become an atomic operation to enforce the order of execution.
 
@@
 The thread scheduler could switch the context after any of these instructions.
-This leads to undefined behavior, when multiple threads try to increment the same variable.
+This leads to unpredictable and non-deterministic behavior, when multiple threads try to increment the same variable.
 To fix this bug pattern, the \lstinline{++} operation also needs to be replaced by an atomic operation that does not allow other threads to access the \lstinline{sum} variable while incrementing.
@@
 
-Assumed, a concurrency bug that rarely manifests, gets deployed into production.
+Assuming that a concurrency bug that rarely manifests, gets deployed into production.
 How can it be found quickly?
@@
 In dynamic code analysis the runtime behavior of a program is observed to detect problematic patterns like un-synchronized variable accesses or deadlocks.
-Assumed, a bug is reported that does not manifest regularly, this technique can help to find the manifestation of the bug more easily during the debugging process.
+Assuming that a bug is reported that does not manifest regularly, this technique can help to find the manifestation of the bug more easily during the debugging process.
 Additionally it could be used to observe deployments in production, so concurrency bugs get reported immediately as they occur.
@@
 \subsection{Record and Replay}
-One popular implementation of this technique is called ``Record and Replay'' where the path of execution of an application is recorded and can later be deterministically replayed.
+One popular dynamic code analysis technique is called ``Record and Replay'' where the path of execution of an application is recorded and can later be deterministically replayed.
 To create a recording that contains all information to replay a program exactly the same way it was executed, the recorder needs to keep track of the schedule of the program as well as all variables that might not be deterministically reproducible.
-Furthermore, any source of non-determinism such as the the interaction with the network or graphics card or any call to external libraries need to be recorded~\cite{lidbury2019sparse}.
+Furthermore, any source of non-determinism such as the interaction with the network or graphics card or any call to external libraries needs to be recorded~\cite{lidbury2019sparse}.
 This creates a lot of computational overhead in CPU time as well as storage.
@@
 
-To track the schedule of a natively compiled program such as Go programs, the record tool has to sit between the host operating system and the application.
+To track the schedule of a native binary application such as Go programs, the record tool has to sit between the host operating system and the application.
 One requirement to use such a tool in production is the ease of deployability which means that the operational overhead should be very small.
 There have been approaches with custom hardware or patched operating system kernels to hijack the scheduler, but these setups are quite complex and not easy to deploy~\cite{o2017engineering}.
-One example for a tool that fulfills the requirements is \emph{rr}~\cite{mozillarr} developed by \emph{Mozilla}.
+One example for a tool that provides record and replay for native binary is \emph{rr}~\cite{mozillarr} developed by \emph{Mozilla}.
+It does not need custom hardware and can run on an unmodified Linux operating system, which makes it easy to use and easy to deploy.
 Their approach is to run the program to observe on a fixed CPU core and track the system calls by using \emph{ptrace}, which is a system call provided by the Linux kernel to inspect and observe processes.
@@
 Ignoring sources of non-determinism does decrease the accuracy in which concurrency bugs can be replayed.
-But for real-world production usage one has to distinguish between the coverage of realistic bug occurrences and the checking of overall correctness of a program.
+But for real-world application one has to distinguish between the coverage of realistic bug occurrences and checking the overall correctness of a program.
 In many cases it is more desirable to observe only one critical region of the code where concurrency bugs are either expected or could lead to serious issues than observing the entire program at once.
@@
 
-Sparse Record and replay provides the ability to make these decisions and therefore adjust the needs to monitor applications in production.
+Sparse record and replay provides the ability to make these decisions and therefore adjust the needs to monitor applications in production.
 Once a bug occurs and is either reported by a user or by the program itself, the developer could just replay the thread schedule to reproduce, find and therefore fix the bug quickly because, ``detecting concurrency bugs is difficult but once detected; correcting them is somehow an easy job''~\cite{tchamgoue2012testing}.
-Additionally, this technique can help to reproduce all kind of concurrency bugs, no matter if they are blocking or non-blocking.
+Additionally, this technique can help to reproduce all kinds of concurrency bugs, no matter if they are blocking or non-blocking.
 For programs that do not heavily depend on parallelism this approach is also very feasible.
@@
 However, for programs that heavily rely on parallelism the overhead to record and store the thread schedule and all necessary parameters can decrease the performance by more than a factor of 30~\cite{o2017engineering}.
-Especially for multi-threaded server applications that expect a high workload and where speed is crucial, this approach is not feasible in production, yet.
+Especially for multi-threaded server applications that expect a high workload and where speed is crucial, this approach is not feasible in production.
 And even if the workload is low but the application is time-sensitive for example because of protocol timeouts, it is not applicable without extending the acceptable timeouts.
@@
 It is build into the Golang compiler tool-chain and can be toggled with a simple compiler flag which makes it very easy to use.
-The tool performs compile-time instrumentation of the source program, in which all (atomic and non-atomic) accesses to potentially shared locations, as well as fence operations, are instrumented with calls into a statically linked run-time library~\cite{lidbury2019sparse}.
+The tool performs compile-time instrumentation of the source program, in which all (atomic and non-atomic) accesses to potentially shared locations, as well as fence operations, are instrumented with calls into a linked run-time library~\cite{lidbury2019sparse}.
 To distinguish between synchronized accesses and data races during runtime, the compiler also has to inject dynamic annotations into the program.
@@
 When the program along with the race detector gets executed, the data race detector observes the program execution as a sequence of events~\cite{serebry2009threadsanitizer}.
-Each memory access is recorded and when receiving events for memory access it can report a potential race based on a \emph{state machine algorithm}~\cite{serebry2011llvm}.
+Each memory access is recorded and when receiving events for memory access it can report a potential race based on a state machine algorithm~\cite{serebry2011llvm}.
 
@@
 Extensive testing has proven to minimize the bugs of software in production~\cite{makinen2014testing}.
-However, traditional testing mostly covers sequential errors and cannot detect concurrency bugs effectively~\cite{lu2008mistakes}.
+However, traditional testing mostly covers errors in sequential programs and cannot detect concurrency bugs effectively~\cite{lu2008mistakes}.
 This lies in the nature of these bugs themselves.
@@
 They make use of a technique called \emph{Delta debugging}, which describes the process of narrowing down the spot where a failure is introduced by going back and forth between working and non-working conditions, in this case working and non-working thread schedules~\cite{zeller2002delta}.
+By identifying the error-introducing thread switch, the corresponding unsafe operation of the program can be found.
 \Cref{fig:testing} shows a modified model of the \emph{DEJAVU} approach, suited for Go programs by using rr instead of of \emph{DEJAVU} for the recording and replay of the thread schedule.
@@
 But for most cases it is not necessary to evaluate all possible thread schedules.
-Lu et al.
-have found in their study of real-world concurrency bugs that almost all concurrency bugs were guaranteed to manifest if a certain partial order between two threads is enforced~\cite{lu2008mistakes}.
+Lu et al.~\cite{lu2008mistakes} have found in their study of real-world concurrency bugs that almost all concurrency bugs were guaranteed to manifest if a certain partial order between two threads is enforced.
 Which means that pairwise testing on concurrent program threads can expose most concurrency bugs, and greatly reduce the testing complexity~\cite{lu2008mistakes}.
@@
 
-Although unfeasible for detecting data races in large code bases, there are multiple implementations of static deadlock-detectors for concurrent programs.
+Although infeasible for detecting data races in large code bases, there are multiple implementations of static deadlock-detectors for concurrent programs.
 One concrete static deadlock detector for the Go programming language is called \emph{Godel Checker}~\cite{godelChecker}.
@@
 One of their main approaches is to encourage developers to synchronize threads via easy to use \emph{message passing}, namely channels.
-A study conducted by Tu, Liu, Song and Zhan~\cite{tu2019go} analyzed popular real-world Go projects and their concurrency bugs.
+However, a study conducted by Tu, Liu, Song and Zhan~\cite{tu2019go} analyzed popular real-world Go projects and their concurrency bugs.
 They came to the surprising conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
-So, synchronization via \emph{message passing} does not seem to be the right solution to decrease concurrency bugs.
 

### 2020-07-15 presentation/slides.tex
Presentation

% !TEX program = lualatex
% !BIB program = biber
\documentclass[english,aspectratio=43,t]{beamer}
% useful options:
%   aspectratio=169

\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\usepackage{listings}
\usepackage[iso]{isodate}
\usepackage[backend=biber]{biblatex}

\addbibresource{../src/references.bib}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single,
    xleftmargin=2em,
    framexleftmargin=1.57em,
    framexrightmargin=-0.5em
}

\lstset{style=mystyle}
% select theme before using any of its features
\usetheme{Comsys}

\title[Cool]{Finding Concurrency Bugs in Production}
\author{Marvin Strangfeld}
\authorextra{Advisor: Julian Büning}
\subtitle{An Analysis of real-world Techniques}
\date{\today}
\location{Aachen}
%\logoextra{logos/rwth_comsys_cropped}

%\begin{titlepicture}
%	\includegraphics[width=3cm]{images/presentation}
%\end{titlepicture}

\begin{document}

\titleframe

\begin{frame}
	\frametitle{Outline}
	\tableofcontents
\end{frame}

\section{Introduction}
\frame\sectionpage

\begin{frame}[c]{The Danger of Concurrency Bugs}
    \centering
    \begin{block}{Therac-25}
        \begin{itemize}
            \item Computer-controlled radiation therapy machine
            \item Two modes of radiation therapy
            \item Depending on software safety checks that only the intended mode is activated
            \item Race condition between the control and interface task
            \item Three people died because of a radiation overdose
        \end{itemize}
    \end{block}
    % \begin{block}{Northeast blackout of 2003}
    %     \begin{itemize}
    %         \item Power outage throughout parts of the Northeastern and Midwestern United States
    %         \item Race condition in the energy management system
    %     \end{itemize}
    % \end{block}
\end{frame}

\note[itemize]{
    \item Some Test
}

\begin{frame}{Motivation}
    \begin{itemize}
        \item<1-> Parallelism is necessary to achieve high performance on multi-core architectures
        \item<2-> Paths of execution (Threads) are executed mostly non-deterministic
        \item<3-> Heisenbugs -- Bugs that only occur within specific thread interleavings
        \item<4-> \Rightarrow{} Concurrency bugs are not easily reproducible
    \end{itemize}
    \begin{exampleblock}{Goals}<5->
        \begin{enumerate}
            \item Finding concurrency bugs effectively with a minimum overhead of work and computational resources
            \item Preventing concurrency bugs to get into production
        \end{enumerate}
    \end{exampleblock}
\end{frame}

\section{Types of Concurrency Bugs}
\frame\sectionpage

\begin{frame}{Taxonomy of Concurrency Bugs}
    \begin{figure}
        \includegraphics[width=\linewidth]{../src/figures/ConcurrencyBugClasses.pdf}
        \caption{A Taxonomy of Concurrency Bugs -- based on\cite{tchamgoue2012testing}}
    \end{figure}
\end{frame}

\begin{frame}[fragile]{Deadlocks}
\begin{itemize}
    \item Circular dependencies between resources / threads block the flow of the program
\end{itemize}
\begin{lstlisting}[language=Go]
func main() {
    data := make(chan int)
    done := make(chan bool)
    data <- 1 // Program already blocks here
    go func() {
        process(<-data)
        done <- true
    }()
    <-done
}
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]{Data Races}
\begin{itemize}
    \item<1-> Multiple threads simultaneously try to access a shared variable where at least one access is a write operation~\cite{serebry2009threadsanitizer}
\end{itemize}
\begin{lstlisting}[language=Go]<1->
func main() {
	data := make(map[string]string)
	done := make(chan bool)
	go func() {
		data["1"] = "a" // First conflicting access.
		done <- true
	}()
	data["2"] = "b" // Second conflicting access.
	<-done
}
\end{lstlisting}
\begin{itemize}
    \item<2-> Semantical data races
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Atomicity and Order Violations}
\begin{itemize}
    \item Interleaving of threads violate the programmer's intention of atomicity and order
\end{itemize}
\begin{lstlisting}[language=Go]
func main() {
    var group sync.WaitGroup
    group.Add(10)
    sum := 0
    for i = 0; i < 10; i++ {
        go func() {
            sum++ // Not atomic
            group.Done()
        }()
    }
    group.Wait()
}
\end{lstlisting}
\end{frame}

\section{Debugging Techniques}
\subsection{Dynamic Code Analysis}
\frame\sectionpage

\begin{frame}{Record and Replay}
    \begin{block}{Idea}<1->
        Recording the exact path of execution so it can be replayed deterministically
    \end{block}
    \begin{itemize}
        \item<2-> Record
        \begin{itemize}
            \item Thread-Schedule of the program
            \item Non-deterministic variables
            \item External sources of non-determinism (e.g. network traffic)
        \end{itemize}
        \item<3-> Sparse recording~\cite{lidbury2019sparse}
        \begin{itemize}
            \item Selectively ignoring unimportant sources of non-determinism
        \end{itemize}
        \item<4-> \emph{rr} by Mozilla~\cite{mozillarr}
        \begin{itemize}
            \item Can be used with native programs written in any language
            \item No instrumentation needed \Rightarrow{} less perturbation
            \item Runs on stock Linux kernels without system modification
            \item Emulates a single-core machine
            \item Overhead depends on the workload
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Data Race Detection}
    \begin{block}{Idea}<1->
        Track program execution to find data races during runtime
    \end{block}
    \begin{itemize}
        \item<2-> \emph{ThreadSanitizer} v2 by Google~\cite{threadSanitizer}
        \begin{itemize}
            \item Build into the Go compiler tool-chain
            \item Compile-time instrumentation of accesses to shared locations
            \item Dynamic annotations to understand the synchronization primitives
            \item Report of potential races based on a state-machine algorithm
        \end{itemize}
        \item<3-> Overhead
        \begin{itemize}
            \item Memory: 5-10x
            \item Execution time: 2-20x
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}[c]{Dynamic Code Analysis}
    \centering
    \begin{exampleblock}{Advantages}
        \begin{itemize}
            \item Easy to use
            \item Could be deployed to production
            \item Can be combined with other debugging tools
        \end{itemize}
    \end{exampleblock}
    \begin{alertblock}{Disadvantages}
        \begin{itemize}
            \item Overhead during execution time
            \item Coverage is quite small
            \item Cannot prevent concurrency bugs to get into production
        \end{itemize}
    \end{alertblock}
\end{frame}

\subsection{Concurrency-aware Testing}
\frame\sectionpage

\begin{frame}{Concurrency-aware Testing}
    \begin{block}{Idea}<1->
        Testing the program for concurrency bugs before it gets deployed into production
    \end{block}
    \begin{itemize}
        \item<2-> Traditional testing mostly covers errors in sequential programs~\cite{lu2008mistakes}
        \item<3-> DEJAVU~\cite{acm2002}
        \begin{itemize}
            \item For the Jalapeno JVM
            \item Using record and replay and delta debugging
            \item Going back and forth between working and failing thread schedules
            \item Narrowing down the failure-inducing thread switch
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}
\begin{figure}
    \includegraphics[height=\textheight]{../src/figures/Concurrency-Testing.pdf}
\end{figure}
\end{frame}

\begin{frame}[c]{Concurrency-aware Testing}
    \centering
    \begin{exampleblock}{Advantages}
        \begin{itemize}
            \item Finding concurrency bugs automatically
            \item Can prevent concurrency bugs to get into production
        \end{itemize}
    \end{exampleblock}
    \begin{alertblock}{Disadvantages}
        \begin{itemize}
            \item Operational overhead for setting up the system
            \item Testing time can grow exponentially with the number of threads
            \item Coverage depends on the quality of the tests (semantical correctness)
        \end{itemize}
    \end{alertblock}
\end{frame}

\subsection{Static Code Analysis}
\frame\sectionpage

\begin{frame}{Static Code Analysis}
    \begin{block}{Idea}<1->
        Finding concurrency bugs during compile time by statically analyzing the program
    \end{block}
    \begin{itemize}
        \item<2-> Translating the source code into mathematical models that can be checked
        \item<3-> Mostly infeasible for detecting data races due to the large number of possible thread interleavings
        \item<4-> \emph{Godel Checker}~\cite{godelChecker}
        \begin{itemize}
            \item Static deadlock detector for Go
            \item Converts program code to single static assignments
            \item These can be analyzed by classic model and termination checkers
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}[c]{Static Code Analysis}
    \centering
    \begin{exampleblock}{Advantages}
        \begin{itemize}
            \item Finding concurrency bugs at compile time
            \item Can prevent concurrency bugs to get into production
        \end{itemize}
    \end{exampleblock}
    \begin{alertblock}{Disadvantages}
        \begin{itemize}
            \item Takes a lot of time to check
            \item Only feasible for programs with little parallelism
        \end{itemize}
    \end{alertblock}
\end{frame}

\section{Conclusion}
\frame\sectionpage

\begin{frame}{Conclusion}
    \begin{itemize}
        \item<1-> Dynamic Code Analysis
        \begin{itemize}
            \item Record and replay can be used to deterministically reproduce bugs
            \item Dynamic data race detection can be used to find data races at runtime
            \item \textbf{Advantage:} Easy to use and easy to deploy
            \item \textbf{Disadvantage:} Large overhead and small coverage
        \end{itemize}
        \item<2-> Concurrency-aware Testing
        \begin{itemize}
            \item Testing different thread schedules for concurrency bugs
            \item \textbf{Advantage:} Prevent concurrency bugs to get into production
            \item \textbf{Disadvantage:} Operational overhead of setting up and maintaining the system and tests
        \end{itemize}
        \item<3-> Static Code Analysis
        \begin{itemize}
            \item Translate code into mathematical models to check correctness
            \item \textbf{Advantage:} Find concurrency bugs at compile time
            \item \textbf{Disadvantage:} Only feasible for applications with little parallelism
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}[noframenumbering,plain,allowframebreaks]{Sources}
    \printbibliography[heading=none]
\end{frame}

\end{document}

### 2020-07-15 examples/replay/main.go

@@ -1,0 +1,22 @@
+package main
+
+import (
+	"fmt"
+	"sync"
+)
+
+var data = []string{"1", "2", "3", "4"}
+
+// Based on "Understanding real-worldconcurrency bugs in go"
+// by T. Tu, X. Liu, L. Song and Y. Zhang
+func main() {
+	var group sync.WaitGroup
+	group.Add(len(data))
+	for _, d := range data {
+		go func(d string) {
+			fmt.Printf("Processing: %s\n", d)
+			defer group.Done()
+		}(d)
+	}
+	group.Wait() // This is causing the deadlock
+}

### 2022-05-15 src/paper-skill.tex
add SKILL entry
Signed-off-by: Marvin Strangfeld <marvin@strangfeld.io>

--- 2020-06-28 src/paper.tex
@@
-% ------------------------------------ %
-% Copyright (C) 2020 Marvin Strangfeld %
-% ------------------------------------ %
-\documentclass[conference]{IEEEtran}
+% !TeX encoding = UTF-8
+% !TeX program = pdflatex
+% !BIB program = biber
 
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
+\documentclass[english]{lni}
+
+\usepackage[
+    backend=biber,
+    style=LNI,
+]{biblatex}
+\addbibresource{references.bib}
+
 \usepackage{algorithmic}
@@
 \usepackage{xcolor}
-\usepackage[hidelinks]{hyperref}
-\usepackage{cleveref}
 \usepackage{float}
@@
 
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 \begin{document}
-
 \title{Finding Concurrency Bugs in Production}
-
-\author{\IEEEauthorblockN{Marvin Alexander Strangfeld}
-\IEEEauthorblockA{
-    \textit{RWTH Aachen University}\\
-    marvin.strangfeld@rwth-aachen.de}
-}
-
+\author[Marvin Strangfeld]
+{Marvin Strangfeld\footnote{RWTH Aachen University, \email{marvin.strangfeld@rwth-aachen.de}}}
+\editor{Gesellschaft für Informatik} % Names of Editors
+\booktitle{SKILL 2022} % Name of book title
+\yearofpublication{2022}
+%%%\lnidoi{18.18420/provided-by-editor-02} % if known
 \maketitle
 
-% ------------------------------------ %
-% ------------- ABSTRACT ------------- %
-% ------------------------------------ %
 \begin{abstract}
@@
 The goal is to find tools that are production-ready and applicable to real-world software projects.
-The techniques evaluated are:
-\begin{itemize}
-    \item Dynamic code analysis in form of record and replay and dynamic data race detection.
-    \item Concurrency-aware testing by evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
-    \item Static code analysis with the methods of sequentialization and model-checking.
-    \item Programming language design patterns to avoid concurrency bugs when writing software.
-\end{itemize}
 \end{abstract}
-
-\begin{IEEEkeywords}
-    [Software Engineering]: Software testing and Debugging
-    [Computing methodologies]: Concurrent programming languages
-\end{IEEEkeywords}
-
-% ------------------------------------ %
-% ----------- INTRODUCTION ----------- %
-% ------------------------------------ %
+\begin{keywords}
+    Software testing and Debugging \and
+    Concurrent programming languages
+\end{keywords}
+%%% Beginn des Artikeltexts
 \section{Introduction}
@@
 The goal of these languages is to ease the usage of parallel programming patterns, so developers can use them more often without writing extensively more code.
-Additionally, they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in \Cref{sct:outlook}.
 One example for such a language is Go which is a statically typed programming language developed by Google, that compiles down to native binaries~\cite{goDocs}.
@@
 Beside the message passing however, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
-Even though Go is designed for programming concurrent programs, this does not mean that Go programs are free of concurrency bugs as we will see in \Cref{sct:outlook}.
+Even though Go is designed for programming concurrent programs, this does not mean that Go programs are free of concurrency bugs.
 Projects written in Go also tend to have more concurrency than projects written in traditional programming languages such as C~\cite{tu2019go}.
@@
 \Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically in a testing environment, where in contrast to a production environment concurrency bugs are explicitly provoked.
-\Cref{sct:static} covers how to detect concurrency bugs with static code analysis.
-And finally, \Cref{sct:outlook} will provide a quick outlook on how the examined tools can be used together as well as other promising technologies that might help avoiding concurrency bugs in the future.
+Finally, \Cref{sct:static} covers how to detect concurrency bugs with static code analysis.
 
@@
 \begin{figure}
-    \includegraphics[width=\linewidth]{figures/ConcurrencyBugClasses.pdf}
+    \centering
+    \includegraphics[width=0.7\linewidth]{figures/ConcurrencyBugClasses.pdf}
     \caption{A Taxonomy of Concurrency Bugs -- based on\cite{tchamgoue2012testing}}
@@
 \begin{figure}
-    \includegraphics[width=\linewidth]{figures/Concurrency-Testing.pdf}
+    \centering
+    \includegraphics[height=0.4\textheight]{figures/Concurrency-Testing.pdf}
     \caption{Automatic concurrency-aware testing for Go programs -- based on\cite{acm2002}}
@@
 % ------------------------------------ %
-% -------------- OUTLOOK ------------- %
-% ------------------------------------ %
-\section{Outlook}
-\label{sct:outlook}
-
-All the techniques shown can find concurrency bugs with different coverage based on the overhead one is willing to pay.
-As seen, there is not a single tool that fulfills all the requirements for an optimal production usage.
-Either the overhead is very big or the coverage is quite small.
-To get the most out of these tools, a combined approach seems to make sense to get the most coverage with the least overhead.
-
-For the goal of having no concurrency bugs at all in a production deployment with nearly no overhead, it would be desirable to have techniques that make it impossible to even write incorrect concurrent programs.
-New languages and language features have been developed for this task.
-Go as presented before is one of those languages that tries to make the use of concurrent patterns safer.
-One of their main approaches is to encourage developers to synchronize threads via easy to use \emph{message passing}, namely channels.
-However, a study conducted by Tu, Liu, Song and Zhan~\cite{tu2019go} analyzed popular real-world Go projects and their concurrency bugs.
-They came to the surprising conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
-
-A promising solution to atomicity and order violation bugs is \emph{software transactional memory} (STM) as proposed by Peyton Jones~\cite{peytonjones2007beautiful}.
-It is an alternative to traditional lock-based synchronization where atomic regions get declared explicitly.
-This way STM is one possibility to ensure that there are no side-effects of atomic operations and no data races.
-Languages like Haskell provide STM by their design of language but Go can also utilize this mechanism by using external libraries.
-Lu, Park, Seo and Zhou~\cite{lu2008mistakes} conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs.
-One of their key finding was that most of the examined non-deadlock concurrency bugs were covered by atomicity-violation or order-violation patterns.
-Their conclusion is that using STM could have avoided about one third of the examined concurrency bugs.
-
-Combining all this with the ongoing trend of using micro-service architectures over big monoliths, these tools could get a lot more useful because the code bases of each service tend to get smaller.
-This means that the overhead of instrumenting, testing or model-checking one service gets smaller in absolute numbers.
-This way teams can decide more easily which applications need to have a higher concurrency bug coverage and which applications can be ignored.
-Record and replay could also be deployed to just some replicas of an instance, reducing the impact on speed whilst collecting useful information to optimize the whole code base.
-
-% ------------------------------------ %
 % ------------ CONCLUSION ------------ %
@@
 
-\bibliographystyle{IEEEtran}
-\bibliography{references}
-
+% \bibliography{references}
+\printbibliography
 \end{document}
